---
title: "The Gap That Didn't Close"
subtitle: "Income, race and US adult obesity, 2011-2024"
author: "Weston T. Nyabeze"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    fig-width: 8
    fig-height: 4.8
    fig-dpi: 200
    embed-resources: true
    smooth-scroll: true
execute:
  warning: false
  message: false
  echo: true
knitr:
  opts_chunk:
    dev: "png"
    dev.args:
      bg: "white"
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(readr); library(dplyr); library(tidyr); library(ggplot2)
  library(stringr); library(forcats); library(purrr); library(tibble)
  library(scales); library(broom); library(knitr); library(kableExtra)
  library(ranger); library(gbm); library(forecast); library(cluster)
  library(plotly)
})
set.seed(20260902)

# Shared palette and theme. Categorical hues are assigned in a fixed order and
# never cycled; ordered variables get a single-hue sequential ramp instead,
# because their job is magnitude rather than identity.
PAL_CAT <- c("#2a78d6", "#eb6834", "#1baf7a", "#eda100",
             "#e87ba4", "#008300", "#4a3aa7", "#e34948")
PAL_SEQ <- c("#86b6ef", "#5598e7", "#3987e5", "#2a78d6",
             "#256abf", "#1c5cab", "#184f95", "#104281")
INK_PRIMARY <- "#0b0b0b"; INK_SECONDARY <- "#52514e"; INK_MUTED <- "#8a8984"
SURFACE <- "#ffffff"; GRID <- "#e8e7e3"

theme_report <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      text = element_text(colour = INK_PRIMARY),
      plot.title = element_text(face = "bold", size = rel(1.12), margin = margin(b = 3)),
      plot.subtitle = element_text(colour = INK_SECONDARY, size = rel(0.95), margin = margin(b = 12)),
      plot.caption = element_text(colour = INK_MUTED, size = rel(0.8), hjust = 0, margin = margin(t = 10)),
      plot.title.position = "plot", plot.caption.position = "plot",
      axis.title = element_text(colour = INK_SECONDARY, size = rel(0.9)),
      axis.text = element_text(colour = INK_SECONDARY, size = rel(0.85)),
      panel.grid.major = element_line(colour = GRID, linewidth = 0.35),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = SURFACE, colour = NA),
      plot.background = element_rect(fill = SURFACE, colour = NA),
      legend.position = "top", legend.justification = "left",
      legend.title = element_blank(),
      legend.text = element_text(colour = INK_SECONDARY, size = rel(0.85)),
      legend.key.height = unit(0.8, "lines"),
      strip.text = element_text(colour = INK_PRIMARY, face = "bold", size = rel(0.9), hjust = 0)
    )
}
theme_set(theme_report())
pct_axis <- function(...) scale_y_continuous(labels = label_number(suffix = "%"), ...)

tbl <- function(x, ...) {
  kbl(x, booktabs = TRUE, ...) |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  full_width = FALSE, position = "left", font_size = 13)
}

# ---------------------------------------------------------------------------
# Every figure in this report is a live plotly chart: hover any mark for its
# underlying values, click a legend entry to hide a series, double-click one to
# isolate it, and drag to zoom. gg_live() is the single conversion point, so
# the charts are authored once in ggplot and every one of them gets the same
# interaction chrome.
# ---------------------------------------------------------------------------
FONT <- "system-ui, -apple-system, 'Segoe UI', Helvetica, Arial, sans-serif"

gg_live <- function(p, tooltip = "text", hovermode = "closest",
                    legend = TRUE, muted = NULL) {
  ttl <- p$labels$title
  sub <- p$labels$subtitle
  has_xlab <- !is.null(p$labels$x)
  p <- p + labs(title = NULL, subtitle = NULL, caption = NULL)

  # Height is left to the chunk's fig-height so that Quarto sizes the
  # container and the widget identically and nothing is clipped.
  w <- ggplotly(p, tooltip = tooltip)

  # Confidence bands, reference lines and backdrop series are drawn for context
  # only: they are dropped from the hover layer and from the legend so that
  # neither is cluttered by marks the reader is not meant to interrogate.
  seen <- character(0)
  w$x$data <- lapply(w$x$data, function(tr) {
    band  <- !is.null(tr$fill) && identical(tr$fill, "toself")
    greyed <- !is.null(muted) && !is.null(tr$line$color) &&
              any(startsWith(tolower(tr$line$color), tolower(muted)))
    if (band || greyed) {
      tr$hoverinfo <- "skip"; tr$showlegend <- FALSE; return(tr)
    }
    if (!is.null(tr$name)) {
      nm <- sub("^\\((.*),\\s*-?\\d+\\)$", "\\1", tr$name)
      tr$name <- nm; tr$legendgroup <- nm
      if (nm %in% seen) tr$showlegend <- FALSE else seen <<- c(seen, nm)
    }
    tr
  })

  # Plotly titles do not wrap, and the report's content column is narrower than
  # a long headline, so the wrapping is done here explicitly.
  wrap_html <- function(x, width) paste(strwrap(x, width = width), collapse = "<br>")

  title_html <- paste0("<b>", wrap_html(ttl, 58), "</b>",
    if (!is.null(sub)) paste0("<br><span style='font-size:12.5px;color:",
                              INK_SECONDARY, "'>", wrap_html(sub, 76), "</span>"))
  n_lines <- lengths(regmatches(title_html, gregexpr("<br>", title_html))) + 1

  w |>
    layout(
      # Title and legend are anchored to the figure container rather than to
      # the plot area, so neither drifts when a long axis label widens the
      # margins, and the legend can never collide with the axis title.
      title = list(text = title_html, x = 0.005, xanchor = "left",
                   xref = "container", y = 0.97, yanchor = "top",
                   yref = "container",
                   font = list(size = 16.5, color = INK_PRIMARY, family = FONT)),
      margin = list(t = 26 + 20 * n_lines, l = 62, r = 26,
                    b = 46 + (if (has_xlab) 24 else 0) + (if (legend) 40 else 0)),
      hovermode = hovermode,
      showlegend = legend,
      legend = list(orientation = "h", x = 0.005, xanchor = "left",
                    y = 0.01, yanchor = "bottom", yref = "container",
                    font = list(size = 12), title = list(text = "")),
      font = list(family = FONT, size = 12, color = INK_SECONDARY),
      paper_bgcolor = SURFACE, plot_bgcolor = SURFACE,
      hoverlabel = list(bgcolor = "white", bordercolor = GRID, align = "left",
                        font = list(family = FONT, size = 12, color = INK_PRIMARY))
    ) |>
    config(displaylogo = FALSE, displayModeBar = "hover",
           modeBarButtonsToRemove = c("select2d", "lasso2d", "autoScale2d",
                                      "hoverClosestCartesian",
                                      "hoverCompareCartesian", "toggleSpikelines"),
           toImageButtonOptions = list(format = "png", scale = 2))
}

comma <- function(x) format(x, big.mark = ",", trim = TRUE)

# GitHub's markdown renderer cannot display an interactive widget, so four of
# the charts below are also written out as static PNGs for the README. They are
# saved from the same ggplot objects that become the live charts, so the two can
# never drift apart.
dir.create("figures", showWarnings = FALSE)
save_png <- function(p, name, caption = NULL, width = 10.2, height = 5.1) {
  ggsave(file.path("figures", paste0(name, ".png")),
         p + labs(caption = caption) + theme_report(base_size = 12),
         width = width, height = height, dpi = 160, bg = SURFACE)
  invisible(p)
}
```

Every chart in this report is live. Hover any point for its underlying values,
click a legend entry to hide that series, double-click one to isolate it, and
drag across a chart to zoom in. The camera icon in the top right of each chart
saves it as an image.

Every number and figure in this report is produced by the code in this document,
run against a single input file. The source `.qmd`, the input extract and the
rendered HTML are in the project repository.

## Introduction and research question {#sec-intro}

**Research question.** How does adult obesity prevalence vary across income levels
and racial and ethnic groups in the United States, and has that gap widened or
narrowed between 2011 and 2024?

The data is the CDC's Behavioral Risk Factor Surveillance System (BRFSS), the
largest continuously running health survey in the world, covering all 50 states,
the District of Columbia and three territories across fourteen survey years. The
question matters because public health budgets are finite: if the obesity gap
between demographic groups is stable, targeting is a permanent requirement rather
than a temporary correction, and knowing which dimension carries the widest gap
tells you where a fixed budget buys the most.

The analysis reaches four conclusions.

1. National adult obesity rose from **27.4% to 33.1%** between 2011 and 2024, a
   5.7 percentage point increase, and the rise is close to linear.
2. The **race and ethnicity gap is roughly 29 percentage points** and has not
   moved in fourteen years. It is about three times the income gap and two and a
   half times the education gap.
3. The income gap **appears** to have collapsed from 9.5 to 2.7 percentage
   points. It did not. The apparent collapse is an artefact of a change to the
   BRFSS income question in 2021, and @sec-break shows the evidence for that
   directly in the data.
4. Nothing in the fourteen-year record or in the forecasts to 2030 suggests any
   of these gaps is closing on its own.

The fourth of those is the finding with the most policy weight. The third is the
one that would most easily have been reported as a finding by an analysis that
did not look for it.

## Data loading and initial inspection {#sec-load}

The source is the CDC's *Nutrition, Physical Activity, and Obesity - Behavioral
Risk Factor Surveillance System* release. The full file carries nine survey
questions; this analysis uses one, `Q036`, "Percent of adults aged 18 years and
older who have obesity". The extract in `data/` has already been filtered to that
question, which is what keeps the repository small enough to be useful.

```{r load}
RAW_PATH <- "data/brfss_obesity_2011_2024.csv"
raw <- read_csv(RAW_PATH, col_types = cols(.default = col_character()), progress = FALSE)

inspection <- tibble(
  Property = c("Rows", "Columns", "Survey years", "Jurisdictions",
               "Stratification variables", "Rows with no published value"),
  Value = c(
    format(nrow(raw), big.mark = ","),
    as.character(ncol(raw)),
    paste0(min(raw$YearStart), "-", max(raw$YearStart)),
    as.character(n_distinct(raw$LocationAbbr)),
    as.character(n_distinct(raw$StratificationCategory1)),
    format(sum(is.na(raw$Data_Value)), big.mark = ",")
  )
)
tbl(inspection, caption = "What the extract contains")
```

Each row is one published estimate: a prevalence rate for one jurisdiction, in one
year, for one level of one stratification variable. The structure of the file
matters more than it first appears, and @sec-structure returns to it.

```{r strata}
strat_overview <- raw |>
  count(StratificationCategory1, name = "Rows") |>
  left_join(raw |> distinct(StratificationCategory1, Stratification1) |>
              count(StratificationCategory1, name = "Levels"),
            by = "StratificationCategory1") |>
  rename(`Stratification variable` = StratificationCategory1) |>
  arrange(desc(Rows))
tbl(strat_overview, caption = "Rows and levels per stratification variable")
```

Fifty-five jurisdictions appear rather than fifty-one: the fifty states, the
District of Columbia, Puerto Rico, Guam, the US Virgin Islands, and a `US` row
carrying the CDC's own national aggregate. That national row is used throughout
for national figures, in preference to averaging state estimates, because it is
weighted to the national population and a simple average of states is not.

Every non-published value in the file carries the same footnote: *"Data not
available because sample size is insufficient or data not reported."* That is
2,482 of 21,560 rows, and @sec-quality treats the pattern in those suppressions
as information rather than as noise.

## Data cleaning and transformation {#sec-clean}

This section is deliberately long. The transformations here are what make
@sec-eda and @sec-model possible at all: in the raw file, income is text, age is
text, education is text, and the geographic coordinate is a string. None of that
can enter a correlation or a model.

### Numeric field fixes

`Sample_Size` arrives comma-formatted (`"1,367"`), so it parses as text and has to
be stripped before it is a number. `Data_Value` is the prevalence estimate and is
empty on suppressed rows.

```{r clean-numeric}
dat <- raw |>
  transmute(
    year         = as.integer(YearStart),
    state_abbr   = LocationAbbr,
    state        = LocationDesc,
    strat_cat    = StratificationCategory1,
    strat_label  = Stratification1,
    strat_id     = StratificationID1,
    obesity_rate = as.numeric(Data_Value),
    ci_low       = as.numeric(Low_Confidence_Limit),
    ci_high      = as.numeric(High_Confidence_Limit),
    sample_size  = as.numeric(str_remove_all(Sample_Size, ",")),
    footnote     = Data_Value_Footnote,
    geolocation  = GeoLocation
  )
```

### Ordinal transformations {#sec-ordinal}

Income, age and education are ordered categories reported as text. Converting them
to numbers is what lets them enter a correlation, a regression slope or a tree
split as a *quantity* rather than as a set of unrelated labels.

Two of these conversions rest on assumptions that cannot be verified from the
data, and both are stated here rather than buried:

- **The top income bracket is open-ended.** "\$75,000 or greater" has no upper
  bound, so any midpoint is an estimate. It is assigned \$100,000, near the
  weighted mean of respondents in that band and consistent with the conventional
  1.3 to 1.4 times lower-bound rule for open-ended income tails.
- **The top age bracket is open-ended.** "65 or older" is assigned 75 years, close
  to the mean age of the US adult population above 65.

@sec-sensitivity tests how much the conclusions move when these two numbers are
changed.

```{r clean-ordinal}
income_midpoint <- c(
  INCLESS15 = 7500, INC1525 = 20000, INC2535 = 30000,
  INC3550 = 42500,  INC5075 = 62500, INC75PLUS = 100000
)
# INCNR ("Data not reported") is left NA on purpose: it is a non-response
# category, not a point on the income scale.

age_midpoint <- c(
  AGEYR1824 = 21.0, AGEYR2534 = 29.5, AGEYR3544 = 39.5,
  AGEYR4554 = 49.5, AGEYR5564 = 59.5, AGEYR65PLUS = 75.0
)

education_rank <- c(EDUHS = 1L, EDUHSGRAD = 2L, EDUCOTEC = 3L, EDUCOGRAD = 4L)

dat <- dat |>
  mutate(
    income_midpoint = unname(income_midpoint[strat_id]),
    age_midpoint    = unname(age_midpoint[strat_id]),
    education_rank  = unname(education_rank[strat_id])
  )
```

### Nominal encoding

Race and ethnicity has no natural order, so it is one-hot encoded rather than
ranked. Imposing a rank on it would let a linear model read a meaningless slope
across the categories. Sex is binary.

```{r clean-nominal}
race_levels <- c("RACEWHT", "RACEBLK", "RACEHIS", "RACEASN",
                 "RACENAA", "RACEHPI", "RACE2PLUS", "RACEOTH")
for (lv in race_levels) {
  dat[[paste0("race_", tolower(sub("RACE", "", lv)))]] <- as.integer(dat$strat_id == lv)
}
dat <- dat |>
  mutate(sex_male = case_when(strat_id == "MALE" ~ 1L, strat_id == "FEMALE" ~ 0L,
                              TRUE ~ NA_integer_))
```

### Geographic transformation

`GeoLocation` is a single string of the form `"(32.840571, -86.631861)"`. It is
split into latitude and longitude, and states are mapped to Census regions, which
collapses 55 jurisdictions into 4 regions plus territories and reduces the
cardinality a model has to cope with.

```{r clean-geo}
dat <- dat |>
  mutate(
    .geo = str_match(geolocation, "\\(\\s*(-?[0-9.]+)\\s*,\\s*(-?[0-9.]+)\\s*\\)"),
    latitude  = as.numeric(.geo[, 2]),
    longitude = as.numeric(.geo[, 3])
  ) |>
  select(-.geo)

northeast <- c("CT","ME","MA","NH","RI","VT","NJ","NY","PA")
midwest   <- c("IL","IN","MI","OH","WI","IA","KS","MN","MO","NE","ND","SD")
south     <- c("DE","DC","FL","GA","MD","NC","SC","VA","WV","AL","KY","MS",
               "TN","AR","LA","OK","TX")
west      <- c("AZ","CO","ID","MT","NV","NM","UT","WY","AK","CA","HI","OR","WA")

dat <- dat |>
  mutate(region = case_when(
    state_abbr %in% northeast ~ "Northeast", state_abbr %in% midwest ~ "Midwest",
    state_abbr %in% south     ~ "South",     state_abbr %in% west    ~ "West",
    state_abbr %in% c("PR","GU","VI") ~ "Territory",
    state_abbr == "US" ~ "National"
  ) |> factor(levels = c("Northeast","Midwest","South","West","Territory","National")))

stopifnot(!any(is.na(dat$region)))
```

### Data quality flag and reliability filter {#sec-quality}

A suppressed row is dropped from every calculation, but the *fact* that it was
suppressed is kept as a binary flag, because where suppression happens is not
random and is itself a finding.

```{r clean-quality}
MIN_SAMPLE <- 50
dat <- dat |>
  mutate(
    suppressed = as.integer(!is.na(footnote) | is.na(obesity_rate)),
    reliable   = as.integer(suppressed == 0 & !is.na(sample_size) & sample_size >= MIN_SAMPLE)
  )

quality <- tibble(
  Stage = c("Rows in the extract", "Suppressed by CDC", "Below the sample-size floor",
            "Reliable rows retained"),
  Rows = c(nrow(dat), sum(dat$suppressed),
           sum(dat$suppressed == 0 & dat$reliable == 0), sum(dat$reliable))
)
tbl(quality, caption = "What the quality filters remove")
```

The sample-size filter removes nothing, which is a result worth stating: the CDC
has already applied a suppression rule at least as strict as a floor of 50. The
filter stays in the pipeline as a guard, so that a future data refresh with a
looser rule would be caught rather than silently absorbed.

Where the suppression falls is the more interesting part.

```{r suppression-by-race}
supp_race <- dat |>
  filter(strat_cat == "Race/Ethnicity") |>
  group_by(`Group` = strat_label) |>
  summarise(`Rows suppressed (%)` = round(100 * mean(suppressed), 1), .groups = "drop") |>
  left_join(
    dat |> filter(strat_cat == "Race/Ethnicity", state_abbr == "US", year == 2024) |>
      transmute(Group = strat_label,
                `Share of 2024 respondents (%)` = round(100 * sample_size / sum(sample_size), 1)),
    by = "Group") |>
  arrange(desc(`Share of 2024 respondents (%)`))
tbl(supp_race, caption = "Suppression is concentrated in the smallest groups")
```

Race and ethnicity accounts for 2,101 of the 2,482 suppressed rows. Nine in ten
state-year estimates for Native Hawaiian and Pacific Islander adults are
suppressed, and roughly a third for Asian and for American Indian and Alaska
Native adults. This is not a defect in the data so much as an arithmetic
consequence of a state-level sample: the groups whose health disparities are
least visible in the published statistics are the groups the survey is least able
to measure. Every state-level result involving race in this report is therefore
built on the three largest groups, and @sec-cluster names the states that drop out
as a result.

### Retaining the original labels

The readable bracket and the numeric version travel together in the same frame,
so charts can be labelled "\$75,000 or greater" while models see 100000. Nothing
downstream has to reverse a transformation to produce a legible axis.

```{r frames}
analysis <- dat |> filter(reliable == 1, state_abbr != "US")   # states and territories
national <- dat |> filter(state_abbr == "US", reliable == 1)   # CDC national aggregate
```

### A structural fact about this data {#sec-structure}

**BRFSS publishes one stratification variable per row.** A row that reports an
income bracket carries no race, no age, no education and no sex. A row that
reports a race carries no income. The variables never co-occur.

This is easy to miss and it changes what can honestly be claimed. A conventional
feature set treating income, race, age, education and sex as simultaneous
predictors of one outcome cannot be built from this file, because no observation
holds more than one of them. Nothing here can say what obesity prevalence is among
low-income Black adults in Alabama; the data reports low-income adults in Alabama
and Black adults in Alabama, separately.

Two consequences run through the rest of the report. @sec-model encodes each
variable alongside an explicit "is this variable present on this row" indicator,
so that a zero means *not applicable* rather than *zero income*. And @sec-discussion
treats the ecological inference limit as a first-order caveat rather than
boilerplate.

## Exploratory analysis {#sec-eda}

### The national baseline

```{r fig-national}
#| label: fig-national
#| fig-cap: "National adult obesity prevalence, 2011-2024. Source: CDC Behavioral Risk Factor Surveillance System."
#| fig-height: 4.47
nat_total <- national |> filter(strat_cat == "Total") |> arrange(year) |>
  select(year, obesity_rate, ci_low, ci_high, sample_size)

nat_change <- with(nat_total, tibble(
  start = obesity_rate[year == 2011], end = obesity_rate[year == 2024],
  abs_change = end - start, rel_change = 100 * (end - start) / start,
  annual_pp = (end - start) / 13))

g <- ggplot(nat_total, aes(year, obesity_rate, group = 1,
       text = sprintf("<b>%d</b><br>%.1f%% of adults<br>95%% CI %.1f to %.1f<br>%s respondents",
                      year, obesity_rate, ci_low, ci_high, comma(sample_size)))) +
  geom_ribbon(aes(ymin = ci_low, ymax = ci_high), fill = PAL_CAT[1], alpha = 0.16) +
  geom_line(colour = PAL_CAT[1], linewidth = 0.9) +
  geom_point(colour = PAL_CAT[1], size = 2.1) +
  scale_x_continuous(breaks = seq(2011, 2024, 2)) +
  scale_y_continuous(labels = label_number(suffix = "%"),
                     breaks = seq(26, 34, 2), limits = c(25, 35)) +
  labs(title = "US adult obesity prevalence, 2011-2024",
       subtitle = "CDC BRFSS national estimate, shaded band is the 95% confidence interval",
       x = NULL, y = NULL)

save_png(g, "readme-national-trend",
         "Source: CDC Behavioral Risk Factor Surveillance System.")
gg_live(g, legend = FALSE)
```

Prevalence rose from `r sprintf("%.1f%%", nat_change$start)` to
`r sprintf("%.1f%%", nat_change$end)`, an increase of
`r sprintf("%.1f", nat_change$abs_change)` percentage points, or
`r sprintf("%.0f%%", nat_change$rel_change)` in relative terms. The average annual
increase is `r sprintf("%.2f", nat_change$annual_pp)` percentage points and the
correlation between year and prevalence is
`r sprintf("%.3f", cor(nat_total$year, nat_total$obesity_rate))`. Confidence
intervals are narrow throughout, because the annual national sample runs from
roughly 350,000 to 470,000 respondents. The 2020 and 2023 dips are within the
noise of a survey conducted partly by telephone during a pandemic; neither breaks
the trend.

### Income

```{r fig-income-lines}
#| label: fig-income-lines
#| fig-cap: "Obesity prevalence by household income bracket. Source: CDC BRFSS national estimates; 'Data not reported' rows excluded."
#| fig-height: 5.75
income_order <- c("Less than $15,000", "$15,000 - $24,999", "$25,000 - $34,999",
                  "$35,000 - $49,999", "$50,000 - $74,999", "$75,000 or greater")

nat_income <- national |>
  filter(strat_cat == "Income", strat_label %in% income_order) |>
  mutate(strat_label = factor(strat_label, levels = income_order))

g <- ggplot(nat_income, aes(year, obesity_rate, colour = strat_label,
       group = strat_label,
       text = sprintf("<b>%s</b><br>%d: %.1f%%<br>%s respondents",
                      strat_label, year, obesity_rate, comma(sample_size)))) +
  geom_line(linewidth = 0.85) + geom_point(size = 1.6) +
  scale_colour_manual(values = rev(PAL_SEQ[c(1,2,3,5,6,8)])) +
  scale_x_continuous(breaks = seq(2011, 2024, 2)) + pct_axis() +
  labs(title = "Obesity prevalence by household income bracket",
       subtitle = "Darker lines are lower-income brackets; income is ordered, so colour encodes magnitude",
       x = NULL, y = NULL)

gg_live(g)
```

The ordering is what you would expect and it holds in every year: lower income,
higher obesity. What is not expected is the jump in the lightest line between
2020 and 2021. Hold that thought until @sec-break.

The transformation in @sec-ordinal makes it possible to plot prevalence against
income as a number rather than as a set of labels.

```{r fig-income-scatter}
#| label: fig-income-scatter
#| fig-cap: "Obesity rate against the transformed income midpoint. Source: CDC BRFSS state-level estimates, 2011-2024; line is an OLS fit with a 95% interval."
#| fig-height: 4.88
inc_scatter <- analysis |> filter(strat_cat == "Income", !is.na(income_midpoint))
income_lm <- lm(obesity_rate ~ income_midpoint, data = inc_scatter)

g <- ggplot(inc_scatter, aes(income_midpoint, obesity_rate,
       text = sprintf("<b>%s, %d</b><br>%s<br>%.1f%% obese<br>%s respondents",
                      state, year, strat_label, obesity_rate, comma(sample_size)))) +
  geom_point(alpha = 0.11, size = 1.1, colour = PAL_CAT[1]) +
  geom_smooth(aes(text = NULL), method = "lm", formula = y ~ x, se = TRUE,
              colour = PAL_CAT[2], fill = PAL_CAT[2], linewidth = 0.9, alpha = 0.2) +
  scale_x_continuous(labels = label_dollar(scale = 1e-3, suffix = "k")) + pct_axis() +
  labs(title = "Obesity rate against the transformed income midpoint",
       subtitle = "One point per state-year-bracket; the transformation is what makes this plot possible",
       x = "Income bracket midpoint", y = NULL)

gg_live(g, legend = FALSE)
```

The points fall into six vertical strips because income is bracketed, not
continuous: the transformation gives the variable a scale, it does not give it
resolution it never had. The fitted slope is
`r sprintf("%.2f", coef(income_lm)[2] * 10000)` percentage points per additional
\$10,000 of bracket midpoint, with a residual spread that is far larger than the
slope. Income is a real but weak predictor at this level of aggregation, which
@sec-model quantifies properly.

### Race and ethnicity

Eight groups is more than one line chart can carry legibly, and the groups are not
equally measurable. The four largest cover 95% of respondents and are almost never
suppressed; the four smallest are volatile enough that year-to-year movement is
mostly sampling noise. The breakdown is split accordingly.

```{r fig-race-big4}
#| label: fig-race-big4
#| fig-cap: "The four largest race and ethnicity groups. Source: CDC BRFSS national estimates. Race is nominal, so it keeps categorical colour rather than a ranked ramp."
#| fig-height: 5.54
nat_race <- national |> filter(strat_cat == "Race/Ethnicity")
big4 <- c("Non-Hispanic Black", "Hispanic", "Non-Hispanic White", "Non-Hispanic Asian")
nat_race4 <- nat_race |> filter(strat_label %in% big4) |>
  mutate(strat_label = factor(strat_label, levels = big4))

g <- ggplot(nat_race4, aes(year, obesity_rate, colour = strat_label,
       group = strat_label,
       text = sprintf("<b>%s</b><br>%d: %.1f%%<br>%s respondents",
                      strat_label, year, obesity_rate, comma(sample_size)))) +
  geom_line(linewidth = 0.95) + geom_point(size = 1.6) +
  scale_colour_manual(values = PAL_CAT[1:4]) +
  scale_x_continuous(breaks = seq(2011, 2024, 2)) +
  pct_axis(limits = c(5, 45)) +
  labs(title = "The four largest race and ethnicity groups, 95% of respondents",
       subtitle = "Every group rose over the period; the ordering never changed, and the spread never closed",
       x = NULL, y = NULL)

gg_live(g)
```

Every group rose. The rank order never changed. The distance between the top line
and the bottom line in 2024 is almost exactly what it was in 2011.

```{r fig-race-panels}
#| label: fig-race-panels
#| fig-cap: "All eight groups, each against the other seven. Source: CDC BRFSS national estimates. Grey lines repeat all eight series in every panel for reference."
#| fig-height: 5.2
small4 <- setdiff(unique(as.character(nat_race$strat_label)), big4)
race_panel <- nat_race |>
  mutate(panel = factor(strat_label, levels = c(big4, small4),
                        labels = str_wrap(c(big4, small4), 17)))
backdrop <- nat_race |> select(year, obesity_rate, g = strat_label)

g <- ggplot(race_panel, aes(year, obesity_rate, group = 1,
       text = sprintf("<b>%s</b><br>%d: %.1f%%<br>%s respondents",
                      strat_label, year, obesity_rate, comma(sample_size)))) +
  geom_line(data = backdrop, aes(group = g, text = NULL), colour = GRID, linewidth = 0.5) +
  geom_line(colour = PAL_CAT[1], linewidth = 0.9) +
  geom_point(colour = PAL_CAT[1], size = 0.9) +
  facet_wrap(~ panel, ncol = 4) +
  scale_x_continuous(breaks = c(2012, 2018, 2024)) + pct_axis() +
  labs(title = "All eight groups, each against the other seven",
       subtitle = "The four smaller groups on the bottom row are frequently suppressed and move erratically",
       x = NULL, y = NULL)

gg_live(g, legend = FALSE, muted = GRID)
```

### Age

The age midpoints from @sec-ordinal turn a set of six labels into a curve, and the
curve has a shape that a categorical treatment would have hidden.

```{r fig-age}
#| label: fig-age
#| fig-cap: "Obesity against age, one line per survey year. Source: CDC BRFSS national estimates. The 65+ midpoint is an assumption; see the sensitivity check."
#| fig-height: 5.75
nat_age <- national |> filter(strat_cat == "Age (years)", !is.na(age_midpoint))
age_lin  <- lm(obesity_rate ~ age_midpoint, data = nat_age)
age_quad <- lm(obesity_rate ~ age_midpoint + I(age_midpoint^2), data = nat_age)

g <- ggplot(nat_age, aes(age_midpoint, obesity_rate,
       colour = factor(year), group = year,
       text = sprintf("<b>%d</b><br>Age %s: %.1f%%<br>%s respondents",
                      year, strat_label, obesity_rate, comma(sample_size)))) +
  geom_line(linewidth = 0.7) + geom_point(size = 1.4) +
  scale_colour_manual(values = colorRampPalette(PAL_SEQ)(14)) +
  scale_x_continuous(breaks = c(21, 29.5, 39.5, 49.5, 59.5, 75),
                     labels = c("18-24","25-34","35-44","45-54","55-64","65+")) +
  pct_axis() +
  labs(title = "Obesity peaks in middle age and falls after 65",
       subtitle = "Each line is one survey year; click a year in the legend to hide it, or double-click to isolate it",
       x = "Age bracket (plotted at its numeric midpoint)", y = NULL)

gg_live(g)
```

A straight line through this relationship explains
`r sprintf("%.0f%%", 100 * summary(age_lin)$r.squared)` of the variance. Adding a
quadratic term takes that to
`r sprintf("%.0f%%", 100 * summary(age_quad)$r.squared)`. The relationship is not
linear and any model that treats it as linear will be wrong at both ends. The
decline after 65 is well documented and is partly survivorship and partly the
weight loss that accompanies illness in older adults, so it should not be read as
older adults being healthier.

### The disparity gaps {#sec-gaps}

The gap for a dimension in a year is the highest group prevalence minus the
lowest. It is a crude measure and a legible one.

```{r gaps}
gap_series <- function(df, cat, keep = NULL) {
  d <- df |> filter(strat_cat == cat)
  if (!is.null(keep)) d <- d |> filter(strat_label %in% keep)
  d |> group_by(year) |>
    summarise(high = max(obesity_rate), low = min(obesity_rate),
              high_label = strat_label[which.max(obesity_rate)],
              low_label  = strat_label[which.min(obesity_rate)],
              gap = high - low, .groups = "drop") |>
    mutate(dimension = cat)
}

gaps <- bind_rows(
  gap_series(national, "Income", keep = income_order),
  gap_series(national, "Race/Ethnicity"),
  gap_series(national, "Education")
) |> mutate(dimension = recode(dimension, "Race/Ethnicity" = "Race / ethnicity"))

gap_summary <- gaps |>
  group_by(Dimension = dimension) |>
  summarise(`Gap 2011` = gap[year == 2011], `Gap 2024` = gap[year == 2024],
            Change = `Gap 2024` - `Gap 2011`,
            `Mean gap` = round(mean(gap), 1),
            `Trend (pp/yr)` = round(coef(lm(gap ~ year))[2], 3), .groups = "drop") |>
  arrange(desc(`Mean gap`))
tbl(gap_summary, caption = "Disparity gaps by dimension, percentage points")
```

```{r fig-gaps}
#| label: fig-gaps
#| fig-cap: "Disparity gaps over time. The dashed line marks the 2021 income question redesign. Source: CDC BRFSS national estimates; the income drop after 2020 is an instrument artefact, not a convergence."
#| fig-height: 5.54
g <- ggplot(gaps, aes(year, gap, colour = dimension, group = dimension,
       text = sprintf("<b>%s, %d</b><br>Gap: %.1f pp<br>%s %.1f%%<br>%s %.1f%%",
                      dimension, year, gap, high_label, high, low_label, low))) +
  geom_vline(xintercept = 2020.5, linetype = "22", colour = INK_MUTED, linewidth = 0.5) +
  geom_line(linewidth = 0.95) + geom_point(size = 1.7) +
  scale_colour_manual(values = PAL_CAT[1:3]) +
  scale_x_continuous(breaks = seq(2011, 2024, 2)) +
  scale_y_continuous(labels = label_number(suffix = " pp"), limits = c(0, 33)) +
  labs(title = "The race and ethnicity gap is three times the income gap and is not closing",
       subtitle = "Highest minus lowest group prevalence within each dimension. Dashed line marks the 2021 income question redesign",
       x = NULL, y = NULL)

save_png(g, "readme-gap-trends",
         "Source: CDC BRFSS national estimates. The income drop after 2020 is an instrument artefact, not a convergence.")
gg_live(g)
```

```{r gap-endpoints}
endpoints <- gaps |> filter(year %in% c(2011, 2024)) |>
  transmute(Dimension = dimension, Year = year,
            `Highest group` = as.character(high_label), `Rate` = high,
            `Lowest group` = as.character(low_label), ` Rate` = low, Gap = gap) |>
  arrange(Dimension, Year)
tbl(endpoints, caption = "Which groups define each gap")
```

The race and ethnicity gap sits near 29 percentage points for all fourteen years.
It is worth being precise about what drives it. The gap runs from Non-Hispanic
Black adults at the top to Non-Hispanic Asian adults at the bottom, and Asian
prevalence is far below every other group.

```{r race-decomp}
race_decomp <- national |> filter(strat_cat == "Race/Ethnicity") |>
  group_by(Year = year) |>
  summarise(
    `All eight groups` = max(obesity_rate) - min(obesity_rate),
    `Excluding Asian adults` =
      max(obesity_rate[strat_label %in% setdiff(big4, "Non-Hispanic Asian")]) -
      min(obesity_rate[strat_label %in% setdiff(big4, "Non-Hispanic Asian")]),
    .groups = "drop") |>
  filter(Year %in% c(2011, 2015, 2020, 2024))
tbl(race_decomp, caption = "The race gap with and without the lowest group")
```

Removing Asian adults from the calculation cuts the gap from about 29 points to
about 10. Both numbers are true and they answer different questions. The 29-point
figure is the full range of outcomes across racial and ethnic groups in the United
States. The 10-point figure is the gap among the three largest groups, and it is
the one that is comparable in scale to the education gap. Reporting only the
larger number without saying what drives it would be misleading; reporting only
the smaller one would drop the group with the best health outcome from the
picture entirely.

### Correlations {#sec-corr}

Correlations are computed within the stratum each variable belongs to. A single
pooled correlation matrix across income, age and education would be built on
non-overlapping cases, for the reason set out in @sec-structure, and would not
mean anything.

```{r correlations}
cor_one <- function(df, var, cat, label) {
  d <- df |> filter(strat_cat == cat, !is.na(.data[[var]]))
  tibble(Variable = label, n = nrow(d),
         Pearson  = round(cor(d[[var]], d$obesity_rate), 3),
         Spearman = round(cor(d[[var]], d$obesity_rate, method = "spearman"), 3),
         `p-value` = format.pval(cor.test(d[[var]], d$obesity_rate)$p.value, digits = 2))
}

correlations <- bind_rows(
  cor_one(analysis, "education_rank",  "Education",   "Education rank"),
  cor_one(analysis, "income_midpoint", "Income",      "Income midpoint"),
  cor_one(analysis, "age_midpoint",    "Age (years)", "Age midpoint"),
  tibble(Variable = "Year (national total)", n = nrow(nat_total),
         Pearson = round(cor(nat_total$year, nat_total$obesity_rate), 3),
         Spearman = round(cor(nat_total$year, nat_total$obesity_rate, method = "spearman"), 3),
         `p-value` = format.pval(cor.test(nat_total$year, nat_total$obesity_rate)$p.value, digits = 2))
) |> arrange(Pearson)
tbl(correlations, caption = "Correlation with obesity prevalence, computed within stratum")
```

Education is the strongest of the three demographic correlations at
`r sprintf("%.2f", correlations$Pearson[correlations$Variable == "Education rank"])`,
ahead of income at
`r sprintf("%.2f", correlations$Pearson[correlations$Variable == "Income midpoint"])`.
Pearson and Spearman agree closely in every case, so the relationships are
monotone and not driven by a handful of extreme values. The age correlation is
positive and moderate, which understates the relationship because it is fitting a
straight line to the curve in @fig-age.

### Regional patterns

```{r fig-region}
#| label: fig-region
#| fig-cap: "Obesity by Census region. Source: CDC BRFSS; territories excluded from the regional cut."
#| fig-height: 5.44
reg_total <- analysis |>
  filter(strat_cat == "Total", region %in% c("Northeast","Midwest","South","West")) |>
  group_by(region, year) |>
  summarise(obesity_rate = weighted.mean(obesity_rate, sample_size),
            n_states = n(), .groups = "drop") |>
  mutate(region = fct_reorder(region, -obesity_rate))

g <- ggplot(reg_total, aes(year, obesity_rate, colour = region, group = region,
       text = sprintf("<b>%s, %d</b><br>%.1f%%<br>%d jurisdictions",
                      region, year, obesity_rate, n_states))) +
  geom_line(linewidth = 0.95) + geom_point(size = 1.6) +
  scale_colour_manual(values = PAL_CAT[1:4]) +
  scale_x_continuous(breaks = seq(2011, 2024, 2)) + pct_axis() +
  labs(title = "Obesity by Census region",
       subtitle = "Sample-size weighted mean of state estimates, collapsing 54 jurisdictions into four regions",
       x = NULL, y = NULL)

gg_live(g)
```

The South is the region usually named in this context, but by 2024 the Midwest is
marginally higher, at
`r sprintf("%.1f%%", reg_total$obesity_rate[reg_total$region == "Midwest" & reg_total$year == 2024])`
against
`r sprintf("%.1f%%", reg_total$obesity_rate[reg_total$region == "South" & reg_total$year == 2024])`.
The Northeast and West track each other closely about five points below.

### A discontinuity in the income series {#sec-break}

Something changed in the income data in 2021, and it is large enough that any
finding about income trends is wrong unless it is accounted for.

`Sample_Size` is usually treated as a reliability weight. It is also, within a
stratification category, a rough census of how respondents were classified. A
sudden change in the composition of a category is therefore a signal that the
survey instrument changed rather than that the population did.

```{r fig-comp-shift}
#| label: fig-comp-shift
#| fig-cap: "Largest year-on-year composition shift in each stratification variable. The 2021 income spike is the question redesign. Source: CDC BRFSS national Sample_Size fields; a shift of this size is an instrument change, not a population change."
#| fig-height: 5.54
composition <- national |>
  group_by(strat_cat, year) |> mutate(share = 100 * sample_size / sum(sample_size)) |> ungroup()

comp_shift <- composition |>
  arrange(strat_cat, strat_label, year) |>
  group_by(strat_cat, strat_label) |> mutate(d_share = share - lag(share)) |> ungroup() |>
  filter(!is.na(d_share)) |>
  group_by(strat_cat, year) |>
  summarise(max_abs_shift = max(abs(d_share)),
            moved = strat_label[which.max(abs(d_share))], .groups = "drop")

g <- ggplot(filter(comp_shift, strat_cat != "Total"),
       aes(year, max_abs_shift, colour = strat_cat, group = strat_cat,
           text = sprintf("<b>%s, %d</b><br>Largest shift: %.1f pp<br>Level that moved: %s",
                          strat_cat, year, max_abs_shift, moved))) +
  geom_line(linewidth = 0.85) + geom_point(size = 1.6) +
  scale_colour_manual(values = PAL_CAT[1:5]) +
  scale_x_continuous(breaks = seq(2012, 2024, 2)) +
  scale_y_continuous(labels = label_number(suffix = " pp")) +
  labs(title = "One variable changed; the other four did not",
       subtitle = "Largest year-on-year shift in the share of respondents falling into any level of each variable",
       x = NULL, y = NULL)

gg_live(g)
```

In 2021, the income composition moved by 22.5 percentage points in a single year.
No other stratification variable moves by more than 2.1 points in any year of the
series, and the income variable's own largest move in any other year is 2.2
points. The 2021 figure is an order of magnitude outside the range of everything
else in the file. This is not a population changing; it is a question changing.

```{r fig-income-comp}
#| label: fig-income-comp
#| fig-cap: "Where the top income bracket went in 2021. Source: CDC BRFSS national estimates; the dashed line marks the break between the 2020 and 2021 instruments."
#| fig-height: 5.54
income_comp <- composition |> filter(strat_cat == "Income") |>
  mutate(grp = case_when(strat_label == "Data not reported"  ~ "Income not reported",
                         strat_label == "$75,000 or greater" ~ "$75,000 or greater",
                         TRUE ~ "All other brackets")) |>
  group_by(grp, year) |> summarise(share = sum(share), .groups = "drop")

g <- ggplot(income_comp, aes(year, share, colour = grp, group = grp,
       text = sprintf("<b>%s</b><br>%d: %.1f%% of respondents", grp, year, share))) +
  geom_vline(xintercept = 2020.5, linetype = "22", colour = INK_MUTED, linewidth = 0.5) +
  geom_line(linewidth = 0.95) + geom_point(size = 1.6) +
  scale_colour_manual(values = PAL_CAT[1:3]) +
  scale_x_continuous(breaks = seq(2011, 2024, 2)) +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  labs(title = "Where the top income bracket went in 2021",
       subtitle = "Share of national respondents by income classification, before and after the instrument change",
       x = NULL, y = NULL)

save_png(g, "readme-income-break",
         "Source: CDC BRFSS national estimates. Dashed line marks the break between the 2020 and 2021 instruments.")
gg_live(g)
```

The mechanism is visible directly. The share of respondents classified as
"\$75,000 or greater" falls from 31.6% in 2020 to 11.6% in 2021, while the share
classified as "income not reported" rises from 16.1% to 38.5%. BRFSS replaced its
income question in 2021 with a version carrying finer brackets above \$75,000, and
respondents in those new higher brackets have no equivalent in the old scheme, so
they land outside the comparable categories.

The consequence for this analysis is direct. After 2020, "\$75,000 or greater" is
no longer the top of the income distribution: it has lost its most affluent and
lowest-obesity members. That is why its measured prevalence jumps from 29.0% to
33.7% in a single year, and that is why the income gap appears to collapse.

```{r income-regimes}
gap_inc <- gaps |> filter(dimension == "Income") |> arrange(year)
income_regimes <- tibble(
  Regime = c("2011-2020, original instrument", "2021-2024, revised instrument"),
  Years = c(10, 4),
  `Gap at start` = c(gap_inc$gap[gap_inc$year == 2011], gap_inc$gap[gap_inc$year == 2021]),
  `Gap at end`   = c(gap_inc$gap[gap_inc$year == 2020], gap_inc$gap[gap_inc$year == 2024]),
  `Trend (pp/yr)` = round(c(coef(lm(gap ~ year, filter(gap_inc, year <= 2020)))[2],
                            coef(lm(gap ~ year, filter(gap_inc, year >= 2021)))[2]), 3)
)
tbl(income_regimes, caption = "The income gap, estimated separately either side of the break")
```

Read across the break, the income gap fell from 9.5 to 2.7 points and the obvious
headline is that America's income-based obesity disparity has almost disappeared.
Read within the original instrument, the gap fell from 9.5 to 8.9 points over ten
years, a decline of `r sprintf("%.2f", abs(coef(lm(gap ~ year, filter(gap_inc, year <= 2020)))[2]))`
points a year that would take roughly sixty years to close. Those are different
findings and only one of them is about incomes.

### Sensitivity to the transformation assumptions {#sec-sensitivity}

@sec-ordinal assigned \$100,000 to the open-ended top income bracket and 75 years
to the open-ended top age bracket. Both are estimates. The check below refits the
two headline relationships across a plausible range for each.

```{r sensitivity}
sens_income <- map_dfr(c(85000, 100000, 120000, 150000), function(v) {
  mid <- income_midpoint; mid["INC75PLUS"] <- v
  d <- analysis |> filter(strat_cat == "Income", strat_id != "INCNR") |>
    mutate(im = unname(mid[strat_id]))
  tibble(`Top bracket midpoint` = label_dollar()(v),
         `Slope per $10k` = round(coef(lm(obesity_rate ~ im, d))[2] * 10000, 3),
         `Pearson r` = round(cor(d$im, d$obesity_rate), 3))
})
tbl(sens_income, caption = "Income relationship under different open-ended midpoints")

sens_age <- map_dfr(c(70, 72, 75, 78), function(v) {
  mid <- age_midpoint; mid["AGEYR65PLUS"] <- v
  d <- national |> filter(strat_cat == "Age (years)") |> mutate(am = unname(mid[strat_id]))
  tibble(`65+ midpoint` = v,
         `Linear R2` = round(summary(lm(obesity_rate ~ am, d))$r.squared, 3),
         `Quadratic R2` = round(summary(lm(obesity_rate ~ am + I(am^2), d))$r.squared, 3))
})
tbl(sens_age, caption = "Age relationship under different open-ended midpoints")
```

The income slope changes in magnitude across the range, because widening the top
bracket's midpoint stretches the x-axis, but the correlation is stable and the
sign and significance never change. The age result is stable throughout: the
quadratic fit beats the linear one by a wide margin at every midpoint tried. No
conclusion in this report depends on either assumption.

## Predictive modelling {#sec-model}

### Feature set

Following @sec-structure, each demographic variable is paired with an indicator
saying whether it is present on that row. Without those indicators a tree would
have to read a zero in `income_midpoint` as a household earning nothing.

```{r model-frame}
race_cols <- grep("^race_", names(analysis), value = TRUE)

model_df <- analysis |>
  mutate(
    has_income = as.integer(!is.na(income_midpoint)),
    has_age = as.integer(!is.na(age_midpoint)),
    has_education = as.integer(!is.na(education_rank)),
    has_sex = as.integer(!is.na(sex_male)),
    has_race = as.integer(strat_cat == "Race/Ethnicity"),
    is_total = as.integer(strat_cat == "Total"),
    income_midpoint = coalesce(income_midpoint, 0),
    age_midpoint    = coalesce(age_midpoint, 0),
    education_rank  = coalesce(as.numeric(education_rank), 0),
    sex_male        = coalesce(as.numeric(sex_male), 0),
    post_2021       = as.integer(year >= 2021)
  ) |>
  select(obesity_rate, year, post_2021, region, income_midpoint, has_income,
         age_midpoint, has_age, education_rank, has_education, sex_male, has_sex,
         all_of(race_cols), has_race, is_total, latitude, longitude,
         state_abbr, strat_cat) |>
  filter(!is.na(latitude), !is.na(longitude))

feature_cols <- setdiff(names(model_df), c("obesity_rate", "state_abbr", "strat_cat"))
```

The frame carries `r format(nrow(model_df), big.mark = ",")` rows and
`r length(feature_cols)` features.

### Two train and test splits

A single random split answers only how well the model interpolates inside the
period it has already seen. The temporal split is the harder question and the one
that matters if the model is ever pointed at a future year.

```{r fit-models}
n_rows <- nrow(model_df)
idx <- sample.int(n_rows, size = floor(0.75 * n_rows))
splits <- list(
  `Random 75/25 rows` = list(train = model_df[idx, ], test = model_df[-idx, ]),
  `Temporal: train to 2021, test 2022-2024` =
    list(train = filter(model_df, year <= 2021), test = filter(model_df, year >= 2022))
)

metrics <- function(actual, pred) {
  r <- actual - pred
  tibble(R2 = 1 - sum(r^2) / sum((actual - mean(actual))^2),
         RMSE = sqrt(mean(r^2)), MAE = mean(abs(r)))
}

fit_all <- function(split, label) {
  tr <- split$train; te <- split$test
  form <- reformulate(feature_cols, response = "obesity_rate")
  m_lm  <- lm(form, data = tr)
  m_rf  <- ranger(form, data = tr, num.trees = 600, mtry = 6, min.node.size = 5,
                  importance = "permutation", seed = 20260902, num.threads = 2)
  m_gbm <- gbm(form, data = tr, distribution = "gaussian", n.trees = 1500,
               interaction.depth = 4, shrinkage = 0.05, bag.fraction = 0.8,
               train.fraction = 0.8, n.minobsinnode = 10, verbose = FALSE)
  best_iter <- gbm.perf(m_gbm, method = "test", plot.it = FALSE)
  preds <- list(`Linear regression` = predict(m_lm, newdata = te),
                `Random forest`     = predict(m_rf, data = te)$predictions,
                `Gradient boosting` = predict(m_gbm, newdata = te, n.trees = best_iter))
  list(results = imap_dfr(preds, ~ metrics(te$obesity_rate, .x) |> mutate(Model = .y)) |>
         mutate(Split = label, n_train = nrow(tr), n_test = nrow(te)),
       lm = m_lm, rf = m_rf, gbm = m_gbm, best_iter = best_iter, test = te, preds = preds)
}

fits <- imap(splits, fit_all)
eval_tbl <- bind_rows(map(fits, "results")) |>
  transmute(Split, Model, R2 = round(R2, 3), RMSE = round(RMSE, 3), MAE = round(MAE, 3),
            `Train rows` = n_train, `Test rows` = n_test) |>
  arrange(Split, desc(R2))
tbl(eval_tbl, caption = "Model comparison on both splits")
```

Gradient boosting wins the random split with an R-squared of
`r eval_tbl$R2[eval_tbl$Split == "Random 75/25 rows"][1]`. The random forest wins
the temporal split. That reversal is informative rather than incidental: boosting
extracts more from patterns inside the observed period and gives some of it back
when asked about years it has not seen, while the forest is more conservative and
holds up better out of sample. Every model loses roughly 10 to 15 points of
R-squared moving from the random split to the temporal one, which is the honest
cost of extrapolating in time.

The linear baseline is far behind both, at
`r eval_tbl$R2[eval_tbl$Model == "Linear regression" & eval_tbl$Split == "Random 75/25 rows"]`
on the random split. Given the curved age relationship in @fig-age and the
interaction between state and demographic group, that is expected.

```{r fig-pred-actual}
#| label: fig-pred-actual
#| fig-cap: "Predicted against actual for the best model on the random split, coloured by which stratification variable the row belongs to. Source: CDC BRFSS; points are state-year-stratum observations."
#| fig-height: 6.37
best_name <- eval_tbl |> filter(Split == "Random 75/25 rows") |> slice(1) |> pull(Model)
fr <- fits[["Random 75/25 rows"]]
pv <- tibble(actual = fr$test$obesity_rate, predicted = fr$preds[[best_name]],
             stratum = fr$test$strat_cat, state = fr$test$state_abbr,
             year = fr$test$year)

g <- ggplot(pv, aes(actual, predicted, colour = stratum,
       text = sprintf("<b>%s, %d</b><br>Stratum: %s<br>Actual %.1f%%, predicted %.1f%%<br>Error %+.1f pp",
                      state, year, stratum, actual, predicted, predicted - actual))) +
  geom_abline(slope = 1, intercept = 0, colour = INK_MUTED, linetype = "22") +
  geom_point(alpha = 0.35, size = 1.1) +
  scale_colour_manual(values = PAL_CAT[1:6]) +
  scale_x_continuous(labels = label_number(suffix = "%")) +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  labs(title = paste0("Predicted against actual, ", tolower(best_name)),
       subtitle = "Held-out 25% of rows. The dashed line is perfect prediction; isolate a stratum by double-clicking its legend entry",
       x = "Actual obesity rate", y = "Predicted")

gg_live(g)
```

```{r resid-stratum}
resid_stratum <- pv |> mutate(r = actual - predicted) |>
  group_by(Stratum = stratum) |>
  summarise(n = n(), MAE = round(mean(abs(r)), 2), Bias = round(mean(r), 2), .groups = "drop") |>
  arrange(desc(MAE))
tbl(resid_stratum, caption = "Where the error is concentrated")
```

The error is not spread evenly. Race and ethnicity rows carry roughly four times
the mean absolute error of `Total` rows. The model is worst precisely where the
underlying estimates are least precise, which is the same small-sample problem
documented in @sec-quality showing up a second time.

### Feature importance {#sec-importance}

```{r fig-importance}
#| label: fig-importance
#| fig-cap: "Grouped permutation importance from the random forest."
#| fig-height: 4.68
imp_raw <- tibble(feature = names(ranger::importance(fr$rf)),
                  importance = as.numeric(ranger::importance(fr$rf)))

group_feature <- function(f) case_when(
  str_starts(f, "race_") | f == "has_race" ~ "Race / ethnicity",
  f %in% c("income_midpoint","has_income") ~ "Income (midpoint)",
  f %in% c("age_midpoint","has_age") ~ "Age (midpoint)",
  f %in% c("education_rank","has_education") ~ "Education (rank)",
  f %in% c("sex_male","has_sex") ~ "Sex",
  f %in% c("latitude","longitude") ~ "Geography (lat/long)",
  f == "region" ~ "Census region",
  f %in% c("year","post_2021") ~ "Year",
  f == "is_total" ~ "Stratum is 'Total'", TRUE ~ f)

imp_grouped <- imp_raw |> mutate(group = group_feature(feature)) |>
  group_by(group) |> summarise(importance = sum(importance), .groups = "drop") |>
  mutate(share = 100 * importance / sum(importance)) |> arrange(desc(importance))

g <- ggplot(mutate(imp_grouped, group = fct_reorder(group, importance)),
       aes(importance, group,
           text = sprintf("<b>%s</b><br>%.1f%% of total importance<br>MSE increase when permuted: %.1f",
                          group, share, importance))) +
  geom_col(fill = PAL_CAT[1], width = 0.62) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.06))) +
  labs(title = "What the random forest actually leans on",
       subtitle = "Permutation importance, with the eight one-hot race columns summed back into one contribution",
       x = "Increase in MSE when the feature is permuted", y = NULL)

gg_live(g, legend = FALSE)
```

Race and ethnicity is the largest single contributor at
`r sprintf("%.0f%%", imp_grouped$share[1])` of total importance, ahead of age at
`r sprintf("%.0f%%", imp_grouped$share[2])`. Income is well down the list. The
gradient boosting model, which computes importance by a different mechanism,
produces the same ordering, which is worth checking because permutation importance
on correlated features can be unstable.

```{r gbm-importance}
gbm_inf <- summary(fr$gbm, n.trees = fr$best_iter, plotit = FALSE) |>
  as_tibble() |> mutate(group = group_feature(as.character(var))) |>
  group_by(Feature = group) |>
  summarise(`Relative influence (%)` = round(sum(rel.inf), 1), .groups = "drop") |>
  arrange(desc(`Relative influence (%)`))
tbl(gbm_inf, caption = "Gradient boosting relative influence, as a cross-check")
```

### A cleaner test of which factor matters most {#sec-incremental}

The ranking above answers the outline's question, but it should be read with one
reservation. Because a race row carries `income_midpoint = 0` and
`has_income = 0`, the model can read the row type off the features, and part of
what looks like the importance of a variable is really the importance of knowing
which kind of row it is.

The comparison below removes that. Within each stratification category
separately, a baseline model is fitted that knows only where and when the
observation was taken. The demographic variable is then added and the increase in
R-squared recorded. Every model is fitted on rows of one kind only, so the
increments are comparable.

```{r fig-incremental}
#| label: fig-incremental
#| fig-cap: "Variance explained by each demographic dimension, state and year held fixed."
#| fig-height: 4.36
incremental_r2 <- function(cat, demo, label) {
  d <- analysis |> filter(strat_cat == cat)
  base <- lm(obesity_rate ~ factor(year) + factor(state_abbr), data = d)
  full <- lm(reformulate(c("factor(year)", "factor(state_abbr)", demo),
                         response = "obesity_rate"), data = d)
  tibble(dimension = label, n = nrow(d),
         r2_baseline = summary(base)$r.squared,
         r2_with_demo = summary(full)$r.squared,
         incremental = summary(full)$r.squared - summary(base)$r.squared)
}

incr <- bind_rows(
  incremental_r2("Age (years)",    "poly(age_midpoint, 2)", "Age"),
  incremental_r2("Race/Ethnicity", "strat_label",           "Race / ethnicity"),
  incremental_r2("Education",      "education_rank",        "Education"),
  incremental_r2("Income",         "income_midpoint",       "Income"),
  incremental_r2("Sex",            "sex_male",              "Sex")
) |> arrange(desc(incremental))

g <- ggplot(mutate(incr, dimension = fct_reorder(dimension, incremental)),
       aes(incremental, dimension,
           text = sprintf("<b>%s</b><br>Adds %.1f%% explained variance<br>Baseline R2 %.3f, with demographic %.3f<br>%s rows",
                          dimension, 100 * incremental, r2_baseline, r2_with_demo,
                          comma(n)))) +
  geom_col(fill = PAL_CAT[1], width = 0.62) +
  scale_x_continuous(labels = label_percent(), expand = expansion(mult = c(0, 0.06))) +
  labs(title = "Which dimension explains the most, holding state and year fixed",
       subtitle = "Extra R-squared when the demographic variable is added to a state-and-year baseline, fitted within each stratum",
       x = "Additional variance explained", y = NULL)

save_png(g + geom_text(aes(label = sprintf("%.0f%%", 100 * incremental)),
                       hjust = -0.2, size = 3.4, colour = INK_SECONDARY),
         "readme-what-matters",
         "Source: CDC BRFSS state-level estimates, 2011-2024. Each row type is modelled separately, so the figures are comparable.",
         height = 4.4)
gg_live(g, legend = FALSE)
```

```{r incremental-table}
tbl(incr |> transmute(Dimension = dimension, n,
                      `Baseline R2` = round(r2_baseline, 3),
                      `With demographic` = round(r2_with_demo, 3),
                      `Increment` = round(incremental, 3)),
    caption = "Incremental variance explained, within stratum")
```

Age and race are close and both are far ahead of the rest. Age explains
`r sprintf("%.0f%%", 100 * incr$incremental[incr$dimension == "Age"])` of the
remaining variance and race
`r sprintf("%.0f%%", 100 * incr$incremental[incr$dimension == "Race / ethnicity"])`;
education `r sprintf("%.0f%%", 100 * incr$incremental[incr$dimension == "Education"])`,
income `r sprintf("%.0f%%", 100 * incr$incremental[incr$dimension == "Income"])`,
and sex essentially nothing. Age and race are doing different work, though. Age is
a life-course pattern that every person passes through, so it is not a disparity
in any policy sense. Race is a difference between people at the same point in
life, and it is the one this analysis set out to measure. Among the dimensions
that constitute disparities, race is first, education second, income third.

## Forecasting to 2030 {#sec-forecast}

Fourteen annual observations is a short series for ARIMA. Rather than lean on one
method, every series below is fitted twice, once with `auto.arima` and once with an
ordinary linear trend, and both are shown. Where the two disagree, the
disagreement is the honest measure of how much the horizon is worth.

```{r forecast-fn}
H <- 6; START <- 2011
forecast_series <- function(values, start_year = START, h = H, label) {
  ts_obj <- ts(values, start = start_year, frequency = 1)
  fit_a <- auto.arima(ts_obj, seasonal = FALSE, stepwise = FALSE,
                      approximation = FALSE, allowdrift = TRUE)
  fc_a  <- forecast(fit_a, h = h, level = 95)
  df <- tibble(year = start_year:(start_year + length(values) - 1), y = values)
  fit_l <- lm(y ~ year, data = df)
  newd <- tibble(year = (max(df$year) + 1):(max(df$year) + h))
  pr <- predict(fit_l, newdata = newd, interval = "prediction", level = 0.95)
  bind_rows(
    tibble(series = label, method = "ARIMA", year = as.numeric(time(fc_a$mean)),
           estimate = as.numeric(fc_a$mean), lo = as.numeric(fc_a$lower[, 1]),
           hi = as.numeric(fc_a$upper[, 1])),
    tibble(series = label, method = "Linear trend", year = newd$year,
           estimate = pr[, "fit"], lo = pr[, "lwr"], hi = pr[, "upr"])
  ) |> mutate(arima_label = as.character(fit_a))
}
```

### The national forecast

```{r fig-forecast-national}
#| label: fig-forecast-national
#| fig-cap: "National adult obesity projected to 2030. Source: CDC BRFSS. Fourteen annual observations; treat the interval, not the point estimate, as the forecast."
#| fig-height: 5.65
fc_nat <- forecast_series(nat_total$obesity_rate, label = "National total")

g <- ggplot() +
  geom_ribbon(data = fc_nat, aes(year, ymin = lo, ymax = hi, fill = method), alpha = 0.16) +
  geom_line(data = nat_total, aes(year, obesity_rate, colour = "Observed", group = 1,
       text = sprintf("<b>%d observed</b><br>%.1f%%<br>95%% CI %.1f to %.1f",
                      year, obesity_rate, ci_low, ci_high)), linewidth = 1) +
  geom_point(data = nat_total, aes(year, obesity_rate, colour = "Observed",
       text = sprintf("<b>%d observed</b><br>%.1f%%<br>95%% CI %.1f to %.1f",
                      year, obesity_rate, ci_low, ci_high)), size = 1.6) +
  geom_line(data = fc_nat, aes(year, estimate, colour = method, group = method,
       text = sprintf("<b>%d, %s forecast</b><br>%.1f%%<br>95%% interval %.1f to %.1f",
                      year, method, estimate, lo, hi)), linewidth = 0.95) +
  geom_point(data = fc_nat, aes(year, estimate, colour = method,
       text = sprintf("<b>%d, %s forecast</b><br>%.1f%%<br>95%% interval %.1f to %.1f",
                      year, method, estimate, lo, hi)), size = 1.5) +
  scale_colour_manual(values = c("Observed" = INK_PRIMARY, "ARIMA" = PAL_CAT[1],
                                 "Linear trend" = PAL_CAT[2])) +
  scale_fill_manual(values = PAL_CAT[1:2], guide = "none") +
  scale_x_continuous(breaks = seq(2011, 2030, 3)) + pct_axis() +
  labs(title = "National adult obesity, projected to 2030",
       subtitle = "Black line is observed BRFSS data; bands are 95% prediction intervals from two independent methods",
       x = NULL, y = NULL)

gg_live(g)
```

`auto.arima` selects `r unique(fc_nat$arima_label)`, which is a random walk with
drift: the best description of the series is that each year adds a constant to the
last. The two methods put 2030 national prevalence at
`r sprintf("%.1f%%", fc_nat$estimate[fc_nat$year == 2030 & fc_nat$method == "ARIMA"])`
and
`r sprintf("%.1f%%", fc_nat$estimate[fc_nat$year == 2030 & fc_nat$method == "Linear trend"])`
respectively, with intervals that overlap heavily. Roughly one in three American
adults today; something closer to 36% by 2030 on current trajectory.

### Group forecasts

The income series is deliberately not forecast past 2020. A model fitted across
the break documented in @sec-break would project the artefact forward as though it
were a trend.

```{r fig-forecast-race}
#| label: fig-forecast-race
#| fig-cap: "Race and ethnicity group forecasts to 2030. Source: CDC BRFSS national estimates. Income groups are excluded; see the note below."
#| fig-height: 5.85
race_groups <- c("Non-Hispanic Black", "Non-Hispanic Asian", "Non-Hispanic White", "Hispanic")
race_hist <- national |> filter(strat_cat == "Race/Ethnicity", strat_label %in% race_groups) |>
  mutate(strat_label = factor(strat_label, levels = race_groups)) |> arrange(strat_label, year)

fc_race <- race_hist |> group_by(strat_label) |>
  group_modify(~ forecast_series(.x$obesity_rate, label = as.character(.y$strat_label))) |>
  ungroup() |> mutate(strat_label = factor(strat_label, levels = race_groups))

g <- ggplot() +
  geom_ribbon(data = filter(fc_race, method == "ARIMA"),
              aes(year, ymin = lo, ymax = hi, fill = strat_label), alpha = 0.15) +
  geom_line(data = race_hist, aes(year, obesity_rate, colour = strat_label,
       group = strat_label,
       text = sprintf("<b>%s</b><br>%d observed: %.1f%%", strat_label, year, obesity_rate)),
       linewidth = 0.95) +
  geom_line(data = filter(fc_race, method == "ARIMA"),
       aes(year, estimate, colour = strat_label, group = strat_label,
           text = sprintf("<b>%s</b><br>%d forecast: %.1f%%<br>95%% interval %.1f to %.1f",
                          strat_label, year, estimate, lo, hi)),
       linewidth = 0.95, linetype = "22") +
  scale_colour_manual(values = PAL_CAT[c(1,4,3,2)]) +
  scale_fill_manual(values = PAL_CAT[c(1,4,3,2)], guide = "none") +
  scale_x_continuous(breaks = seq(2011, 2030, 3)) + pct_axis() +
  labs(title = "Group forecasts to 2030: the top and the bottom do not converge",
       subtitle = "Solid lines observed, dashed lines ARIMA central forecast, shaded 95% prediction intervals",
       x = NULL, y = NULL)

gg_live(g)
```

### The gap projection

The gaps are forecast directly as series rather than as the difference of two
separate forecasts. Differencing two forecasts would require an assumption about
how their errors correlate; forecasting the gap itself does not.

```{r fig-forecast-gaps}
#| label: fig-forecast-gaps
#| fig-cap: "Gap projections to 2030. Source: CDC BRFSS national estimates; dashed lines are ARIMA central forecasts."
#| fig-height: 5.65
gap_race <- gaps |> filter(dimension == "Race / ethnicity") |> arrange(year)
gap_edu  <- gaps |> filter(dimension == "Education") |> arrange(year)
fc_gap <- bind_rows(forecast_series(gap_race$gap, label = "Race / ethnicity gap"),
                    forecast_series(gap_edu$gap,  label = "Education gap"))
gap_hist <- bind_rows(transmute(gap_race, year, gap, series = "Race / ethnicity gap"),
                      transmute(gap_edu,  year, gap, series = "Education gap"))

g <- ggplot() +
  geom_ribbon(data = filter(fc_gap, method == "ARIMA"),
              aes(year, ymin = lo, ymax = hi, fill = series), alpha = 0.15) +
  geom_line(data = gap_hist, aes(year, gap, colour = series, group = series,
       text = sprintf("<b>%s</b><br>%d observed: %.1f pp", series, year, gap)),
       linewidth = 0.95) +
  geom_line(data = filter(fc_gap, method == "ARIMA"),
       aes(year, estimate, colour = series, group = series,
           text = sprintf("<b>%s</b><br>%d forecast: %.1f pp<br>95%% interval %.1f to %.1f",
                          series, year, estimate, lo, hi)),
       linewidth = 0.95, linetype = "22") +
  scale_colour_manual(values = PAL_CAT[c(2,1)]) +
  scale_fill_manual(values = PAL_CAT[c(2,1)], guide = "none") +
  scale_x_continuous(breaks = seq(2011, 2030, 3)) +
  scale_y_continuous(labels = label_number(suffix = " pp"), limits = c(0, 40)) +
  labs(title = "Neither gap is projected to close by 2030",
       subtitle = "Gaps forecast directly as series, so the intervals need no assumption about correlated errors",
       x = NULL, y = NULL)

gg_live(g)
```

```{r forecast-summary}
fc_summary <- bind_rows(fc_nat, fc_gap, fc_race) |>
  filter(year %in% c(2027, 2030)) |>
  transmute(Series = series, Method = method, Year = year,
            Estimate = round(estimate, 1), Low = round(lo, 1), High = round(hi, 1)) |>
  arrange(Series, Method, Year)
tbl(fc_summary, caption = "Forecast summary for 2027 and 2030")
```

For both gaps `auto.arima` selects a model with no trend term, so the central
forecast is a flat line at the historical mean and the prediction interval is the
historical variability. The linear trend model, which is free to move, projects
the race gap slightly *upward* to
`r sprintf("%.1f", fc_gap$estimate[fc_gap$series == "Race / ethnicity gap" & fc_gap$method == "Linear trend" & fc_gap$year == 2030])`
points by 2030. Neither method produces a scenario in which the gap closes.

### How much the break costs the forecast

```{r income-forecast-cost}
fc_inc_full <- forecast_series(gap_inc$gap, label = "Income gap, full 2011-2024 series")
fc_inc_pre  <- forecast_series(gap_inc$gap[gap_inc$year <= 2020], h = 10,
                               label = "Income gap, 2011-2020 only")
tbl(bind_rows(fc_inc_full, fc_inc_pre) |>
      filter(year == 2030, method == "ARIMA") |>
      transmute(Series = series, `2030 estimate` = round(estimate, 1),
                Low = round(lo, 1), High = round(hi, 1)),
    caption = "The same forecast, fitted across the break and inside it")
```

Fitted across the break, the 2030 income gap comes out at 2.7 points with a
prediction interval whose lower bound is negative. A gap defined as a maximum
minus a minimum cannot be negative, so the interval is telling you the model does
not fit. Fitted on the pre-break window alone, the same procedure gives 9.5
points. The two answers differ by a factor of three and only one of them comes
from a model whose assumptions the data supports.

### A backtest

```{r backtest}
backtest <- function(values, label) {
  n <- length(values); k <- 3
  te <- values[(n - k + 1):n]
  forecast_series(values[1:(n - k)], h = k, label = label) |>
    group_by(Method = method) |>
    summarise(MAE = round(mean(abs(estimate - te)), 2),
              RMSE = round(sqrt(mean((estimate - te)^2)), 2), .groups = "drop") |>
    mutate(Series = label)
}
tbl(bind_rows(backtest(nat_total$obesity_rate, "National total"),
              backtest(gap_race$gap, "Race / ethnicity gap"),
              backtest(gap_edu$gap, "Education gap")) |>
      select(Series, Method, MAE, RMSE),
    caption = "Holding out the last three years: how the two methods would have done")
```

The linear trend beats ARIMA on every series. With fourteen observations, the
model selection machinery in `auto.arima` has very little to work with, and the
simpler method wins. Where the two disagree in the tables above, the linear trend
has the better track record on this data.

## Clustering states into disparity profiles {#sec-cluster}

The outline lists region as a clustering feature. It is used here to *validate*
the clusters rather than to build them. k-means minimises Euclidean distance, and
a one-hot region column would contribute an arbitrary distance of root two between
any two regions, a quantity with no meaning. Building the clusters from continuous
features only and then asking whether they line up with geography is both cleaner
and a genuine test of whether they mean anything.

```{r cluster-features}
RECENT <- 2022:2024   # one instrument regime, so the income gap is comparable
BIG3 <- c("Non-Hispanic White", "Non-Hispanic Black", "Hispanic")

overall <- analysis |> filter(strat_cat == "Total", year %in% RECENT) |>
  group_by(state_abbr, state, region) |>
  summarise(obesity = mean(obesity_rate), n_years = n(), .groups = "drop") |>
  filter(n_years == length(RECENT))

trend <- analysis |> filter(strat_cat == "Total") |> group_by(state_abbr) |>
  filter(n() >= 10) |> summarise(trend_pp = coef(lm(obesity_rate ~ year))[2], .groups = "drop")

income_gap <- analysis |>
  filter(strat_cat == "Income", !is.na(income_midpoint), year %in% RECENT) |>
  group_by(state_abbr, year) |> filter(n() == 6) |>
  summarise(gap = max(obesity_rate) - min(obesity_rate), .groups = "drop") |>
  group_by(state_abbr) |> summarise(income_gap = mean(gap), .groups = "drop")

race_gap <- analysis |>
  filter(strat_cat == "Race/Ethnicity", strat_label %in% BIG3, year %in% RECENT) |>
  group_by(state_abbr, year) |> filter(n() == 3) |>
  summarise(gap = max(obesity_rate) - min(obesity_rate), .groups = "drop") |>
  group_by(state_abbr) |> summarise(race_gap = mean(gap), n_yr = n(), .groups = "drop") |>
  filter(n_yr >= 2) |> select(-n_yr)

states <- overall |> inner_join(trend, by = "state_abbr") |>
  inner_join(income_gap, by = "state_abbr") |> inner_join(race_gap, by = "state_abbr") |>
  select(state_abbr, state, region, obesity, income_gap, race_gap, trend_pp)

scaled <- scale(select(states, obesity, income_gap, race_gap, trend_pp))
```

`r nrow(states)` jurisdictions survive the coverage requirements. The ones that
drop out do so because they lack a full three-group race read in at least two of
the three recent years, and they are disproportionately states with small Black or
Hispanic populations. That is a selection effect worth naming: this section
describes disparity profiles among states diverse enough to have measurable ones.

```{r fig-k-selection}
#| label: fig-k-selection
#| fig-cap: "Choosing the number of clusters."
#| fig-height: 4.26
ks <- 2:8; dmat <- dist(scaled)
k_diag <- map_dfr(ks, function(k) {
  km <- kmeans(scaled, centers = k, nstart = 100, iter.max = 200)
  tibble(k = k, within_ss = km$tot.withinss,
         silhouette = mean(silhouette(km$cluster, dmat)[, 3]), smallest = min(km$size))
})
# Silhouette alone picks a solution containing a one-state cluster, which labels
# an outlier rather than describing a profile. Require at least three states.
k_best <- k_diag |> filter(smallest >= 3) |> slice_max(silhouette, n = 1) |> pull(k)

g <- ggplot(k_diag |> select(k, within_ss, silhouette, smallest) |>
         pivot_longer(c(within_ss, silhouette), names_to = "metric", values_to = "value") |>
         mutate(metric = factor(recode(metric,
                  within_ss = "Total within-cluster sum of squares",
                  silhouette = "Mean silhouette width"),
                levels = c("Total within-cluster sum of squares", "Mean silhouette width"))),
       aes(k, value, group = 1,
           text = sprintf("<b>k = %d</b><br>%s: %.3f<br>Smallest cluster: %d states",
                          k, metric, value, smallest))) +
  geom_line(colour = PAL_CAT[1], linewidth = 0.9) + geom_point(colour = PAL_CAT[1], size = 2.2) +
  facet_wrap(~ metric, scales = "free_y") + scale_x_continuous(breaks = ks) +
  labs(title = "The silhouette never rises far, so no k is strongly supported",
       subtitle = paste0("Left, where adding a cluster stops buying much; right, how well separated the clusters are. Selected k = ", k_best),
       x = "Number of clusters", y = NULL)

gg_live(g, legend = FALSE)
```

The silhouette width never exceeds
`r sprintf("%.2f", max(k_diag$silhouette))` at any k tried. That is a weak result
and it is the honest headline of this section: **states lie on a continuum of
disparity profiles rather than falling into distinct types.** A silhouette in this
range means the clusters are convenient labels, not discovered structure. Reporting
them as discovered types would be the mistake this section is designed to avoid.

```{r cluster-fit}
km <- kmeans(scaled, centers = k_best, nstart = 100, iter.max = 200)
states$cluster <- factor(km$cluster)

profile <- states |> group_by(cluster) |>
  summarise(n = n(), obesity = mean(obesity), income_gap = mean(income_gap),
            race_gap = mean(race_gap), trend_pp = mean(trend_pp), .groups = "drop") |>
  arrange(desc(obesity))

LADDER <- list(
  obesity = list(`2` = c("Higher","Lower"), `3` = c("Highest","Middle","Lowest"),
                 `4` = c("Highest","High","Lower","Lowest")),
  gap = list(`2` = c("wider","narrower"), `3` = c("widest","middling","narrowest"),
             `4` = c("widest","wide","narrow","narrowest")))
label_clusters <- function(prof) {
  k <- nrow(prof)
  paste0(LADDER$obesity[[as.character(k)]][rank(-prof$obesity, ties.method = "first")],
         " obesity, ",
         LADDER$gap[[as.character(k)]][rank(-prof$race_gap, ties.method = "first")],
         " race gap")
}
profile$label <- label_clusters(profile)
states <- states |> left_join(select(profile, cluster, label), by = "cluster") |>
  mutate(label = factor(label, levels = profile$label))

tbl(profile |> transmute(Profile = label, States = n,
                         `Obesity (%)` = round(obesity, 1),
                         `Income gap (pp)` = round(income_gap, 1),
                         `Race gap (pp)` = round(race_gap, 1),
                         `Trend (pp/yr)` = round(trend_pp, 2)),
    caption = "Cluster profiles")
```

```{r fig-clusters}
#| label: fig-clusters
#| fig-cap: "State disparity profiles. Source: CDC BRFSS, 2022-2024 averages. Clusters were built without any geographic input."
#| fig-height: 6.58
g <- ggplot(states, aes(race_gap, obesity, colour = label,
       text = sprintf("<b>%s</b> (%s)<br>Obesity %.1f%%<br>Race gap %.1f pp<br>Income gap %.1f pp<br>Trend %+.2f pp/yr<br>Profile: %s",
                      state, region, obesity, race_gap, income_gap, trend_pp, label))) +
  geom_point(size = 3, alpha = 0.85) +
  geom_text(aes(label = state_abbr, text = NULL), size = 2.4, vjust = -1.15,
            show.legend = FALSE, colour = INK_SECONDARY) +
  scale_colour_manual(values = PAL_CAT[1:k_best]) +
  scale_x_continuous(labels = label_number(suffix = " pp")) + pct_axis() +
  labs(title = "State disparity profiles",
       subtitle = paste0("k-means on overall rate, income gap, race gap and 14-year trend; k = ", k_best,
                         ". Hover a state for all four features"),
       x = "Race gap, widest minus narrowest of the three largest groups",
       y = "Overall obesity rate")

gg_live(g)
```

```{r fig-cluster-regions}
#| label: fig-cluster-regions
#| fig-cap: "Census region composition of each cluster. Source: CDC BRFSS. Region was excluded from the clustering features."
#| fig-height: 4.5
region_x <- states |> filter(region != "Territory") |> count(label, region) |>
  group_by(label) |> mutate(share = 100 * n / sum(n)) |> ungroup()
chisq <- suppressWarnings(chisq.test(table(
  droplevels(states$region[states$region != "Territory"]),
  states$label[states$region != "Territory"])))

g <- ggplot(region_x, aes(share, fct_rev(label), fill = region,
       text = sprintf("<b>%s</b><br>%s: %d states (%.0f%% of the cluster)",
                      label, region, n, share))) +
  geom_col(width = 0.62) + scale_fill_manual(values = PAL_CAT[1:4]) +
  scale_x_continuous(labels = label_number(suffix = "%"), expand = expansion(mult = c(0, 0.02))) +
  labs(title = "The clusters recover geography without being told about it",
       subtitle = paste0("Census region composition of each cluster. Chi-squared test of independence: p = ",
                         format.pval(chisq$p.value, digits = 3)),
       x = NULL, y = NULL)

gg_live(g)
```

Region was never given to the algorithm, yet the two clusters split along regional
lines at p = `r format.pval(chisq$p.value, digits = 3)`. The higher-obesity,
narrower-gap cluster is overwhelmingly Midwest and South; the lower-obesity,
wider-gap cluster is Northeast and West. That is a real validation: whatever weak
structure exists in these four features tracks something geographic, which is what
you would expect if the drivers are food environment, urban density and state
policy rather than anything the features name directly.

```{r cluster-members}
tbl(states |> group_by(Profile = label) |>
      summarise(n = n(), Jurisdictions = paste(sort(state_abbr), collapse = ", "),
                .groups = "drop"),
    caption = "Cluster membership")
```

A finer split is worth showing, with the caveat attached rather than buried.

```{r fig-clusters-k4}
#| label: fig-clusters-k4
#| fig-cap: "A four-way split, offered as description rather than as a finding. Source: CDC BRFSS, 2022-2024 averages."
#| fig-height: 6.79
km4 <- kmeans(scaled, centers = 4, nstart = 100, iter.max = 200)
states$cluster4 <- factor(km4$cluster)
profile4 <- states |> group_by(cluster4) |>
  summarise(n = n(), obesity = mean(obesity), income_gap = mean(income_gap),
            race_gap = mean(race_gap), trend_pp = mean(trend_pp), .groups = "drop") |>
  arrange(desc(obesity))
profile4$label4 <- label_clusters(rename(profile4, cluster = cluster4))
states <- states |> left_join(select(profile4, cluster4, label4), by = "cluster4") |>
  mutate(label4 = factor(label4, levels = profile4$label4))

g <- ggplot(states, aes(race_gap, obesity, colour = label4,
       text = sprintf("<b>%s</b> (%s)<br>Obesity %.1f%%<br>Race gap %.1f pp<br>Income gap %.1f pp<br>Trend %+.2f pp/yr<br>Profile: %s",
                      state, region, obesity, race_gap, income_gap, trend_pp, label4))) +
  geom_point(size = 3, alpha = 0.85) +
  geom_text(aes(label = state_abbr, text = NULL), size = 2.4, vjust = -1.15,
            show.legend = FALSE, colour = INK_SECONDARY) +
  scale_colour_manual(values = PAL_CAT[1:4]) +
  scale_x_continuous(labels = label_number(suffix = " pp")) + pct_axis() +
  labs(title = "A four-way split, offered as description rather than as a finding",
       subtitle = "Same features, k = 4. The silhouette barely improves on k = 2, so read these as labels on a continuum",
       x = "Race gap, widest minus narrowest of the three largest groups",
       y = "Overall obesity rate")

gg_live(g)
```

```{r cluster4-table}
tbl(profile4 |> transmute(Profile = label4, States = n,
                          `Obesity (%)` = round(obesity, 1),
                          `Income gap (pp)` = round(income_gap, 1),
                          `Race gap (pp)` = round(race_gap, 1),
                          `Trend (pp/yr)` = round(trend_pp, 2)),
    caption = "The four-way profiles")
```

The four-way version separates two groups that a policymaker would treat
differently. A high-obesity, narrow-gap state has a population-wide problem and
should be reached with population-wide measures. A high-obesity, wide-gap state
has a targeting problem, where the same total burden is distributed very unevenly
and universal programmes will reach the wrong people first.

## Discussion {#sec-discussion}

**Race is the widest disparity and it is not moving.** At roughly 29 percentage
points nationally, the race and ethnicity gap is three times the income gap and
two and a half times the education gap, and fourteen years of data show no trend
in it. @sec-importance and @sec-incremental agree from two different directions:
of the dimensions that constitute a social disparity, race carries the most
explanatory weight. The caveat in @sec-gaps stands, that the 29-point figure is
driven by Asian adults at the low end and the gap among the three largest groups
is closer to 10 points, and both numbers belong in any honest summary.

**The income convergence is not real.** This is the finding an analysis could most
easily get wrong. Read at face value, the income gap fell by 72% between 2011 and
2024 and the natural conclusion is that income-based disparity in obesity has
nearly disappeared. @sec-break shows that the 2021 BRFSS income question redesign
moved twenty percentage points of respondents out of the top bracket, and that the
top bracket's measured prevalence jumped 4.7 points in the same year as a
consequence. Within a single instrument regime the gap is roughly flat. The
diagnostic that caught this is simple and general: watch the composition of a
stratification variable over time, and treat a large jump as an instrument change
until proven otherwise.

**Forecasts extend the trend but do not close the gaps.** National prevalence is
projected to reach roughly 36% by 2030 on either method. Neither gap is projected
to close. Both projections rest on fourteen observations and the backtest in
@sec-forecast shows the simpler linear method outperforming ARIMA on this data, so
the intervals rather than the point estimates are what should be quoted.

**Ecological inference is the binding limitation.** Every result here concerns
groups within states, not individuals. Nothing in this data can say what obesity
prevalence is among low-income Black adults in Alabama, because @sec-structure's
structural fact means the file never reports two demographic variables together.
The apparent independence of income and race in @sec-importance may therefore
reflect the data's structure rather than the world: income and race are correlated
in the US population, and an analysis that cannot see them together cannot
separate their effects. This is the single strongest argument for moving to BRFSS
individual-level microdata for any follow-up.

**The transformation assumptions are stated and tested.** The \$100,000 midpoint
for the open-ended top income bracket and the 75-year midpoint for "65 or older"
are estimates, not observations. @sec-sensitivity refits both relationships across
a plausible range and no conclusion changes.

**Other caveats.** BRFSS height and weight are self-reported, and self-reported
BMI is known to understate obesity, so every level in this report is likely a
floor rather than a point estimate. Suppression is concentrated in the smallest
racial and ethnic groups, so state-level race results are systematically thinner
for the groups where disparities may be largest. Survey mode and weighting shifted
during 2020 and 2021. And state, income and race distributions are mutually
confounded: a state-level association between race and obesity partly reflects
which states particular populations live in.

## Conclusion and recommendations {#sec-conclusion}

The gap that matters most in US adult obesity is racial, it is about 29 percentage
points wide, and it has not narrowed in fourteen years. The income gap, which a
face-value reading would report as having nearly closed, has not closed either;
the appearance of convergence is an artefact of a survey instrument change in 2021.
Both models and both forecasting methods agree that current trajectories carry
national prevalence to roughly 36% by 2030 with the gaps intact.

Three practical implications follow.

1. **Target by race and geography together, not by income alone.** Income is the
   weakest of the three social dimensions measured here, explaining
   `r sprintf("%.0f%%", 100 * incr$incremental[incr$dimension == "Income"])` of
   within-stratum variance against
   `r sprintf("%.0f%%", 100 * incr$incremental[incr$dimension == "Race / ethnicity"])`
   for race. The state clusters in @sec-cluster separate places with a
   population-wide problem from places with a distributional one, and those two
   situations call for different programmes.
2. **Treat the flat race gap as evidence that current approaches are not working
   on it.** Fourteen years of rising prevalence with a constant gap means every
   group is getting worse at roughly the same rate. Interventions that lift all
   groups equally will not close it; only something that acts differentially will.
3. **Fix the income series before using it for policy.** Any analysis of BRFSS
   income trends spanning 2021 needs to handle the break explicitly. The
   composition diagnostic in @sec-break takes about ten lines of code and should be
   standard practice on any long survey panel.

**Next steps.** Three extensions would materially strengthen this work. BRFSS
individual-level microdata would remove the ecological inference limit and allow
income and race to be modelled jointly, which is the question this data cannot
answer. External predictors such as food access, healthcare access and physical
activity infrastructure would move the analysis from description toward
explanation; the same BRFSS release carries physical activity and nutrition
questions that were not used here. And a county-level or census-tract analysis
would replace the 55-jurisdiction geography with something fine enough to site an
actual intervention.

---

**Data source.** Centers for Disease Control and Prevention, *Nutrition, Physical
Activity, and Obesity - Behavioral Risk Factor Surveillance System*, accessed via
[data.cdc.gov](https://data.cdc.gov/Nutrition-Physical-Activity-and-Obesity/Nutrition-Physical-Activity-and-Obesity-Behavioral/hn4x-zwk7/about_data).
