---
title: "The Empty Chair"
subtitle: "Quantifying and Reducing the Cost of Patient No-Shows Across a UK Private Clinic Network"
author: "Weston Nyabeze | MS Business Analytics and AI, Kogod School of Business, American University"
abstract: |
  A ten-page executive edition of the full technical no-show analysis. It states the size of the
  loss, the three operational factors that drive it, how the predictive model was built and
  validated, and a costed four-part action plan. The full technical report, with the complete data
  audit, four-model bake-off and 41 supporting exhibits, is the companion document to this one.
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    toc-location: left
    number-sections: false
    fig-width: 8.5
    fig-height: 4.6
    fig-align: center
    fig-cap-location: bottom
    tbl-cap-location: top
    code-fold: true
    code-summary: "Show the code behind this exhibit"
    code-tools: true
    code-overflow: wrap
    smooth-scroll: true
    anchor-sections: true
    link-external-newwindow: true
    embed-resources: true
    css: styles.css
    include-in-header:
      text: |
        <link rel="preconnect" href="https://fonts.googleapis.com">
        <link rel="preconnect" href="https://fonts.gstatic.com" crossorigin>
        <link href="https://fonts.googleapis.com/css2?family=Source+Serif+4:opsz,wght@8..60,400;8..60,600;8..60,700&family=Inter:wght@400;500;600;700&family=IBM+Plex+Mono:wght@400;500&display=swap" rel="stylesheet">
  docx:
    toc: false
    fig-width: 6.6
    fig-height: 3.7
    fig-align: center
    fig-cap-location: bottom
execute:
  warning: false
  message: false
  echo: true
  cache: false
knitr:
  opts_chunk:
    dev: "png"
    dpi: 200
    fig.retina: 2
editor: source
---

```{r}
#| label: setup
#| include: false

# ---------------------------------------------------------------------------
# 0. ENVIRONMENT
# ---------------------------------------------------------------------------
required <- c("data.table", "ggplot2", "scales", "glmnet", "ranger", "pROC",
              "knitr", "kableExtra", "patchwork", "plotly", "htmltools")
missing_pkgs <- required[!sapply(required, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs)) {
  stop("Install the following packages before rendering: ",
       paste(missing_pkgs, collapse = ", "))
}

suppressPackageStartupMessages({
  library(data.table); library(ggplot2); library(scales)
  library(glmnet);     library(ranger);  library(pROC)
  library(knitr);      library(kableExtra)
  library(patchwork);  library(plotly);  library(htmltools)
})

SEED <- 1234568                          # one seed governs every stochastic step
set.seed(SEED)                           # change it here and the whole report follows
options(scipen = 999, digits = 4)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

if (!grepl("utf-?8", Sys.getlocale("LC_CTYPE"), ignore.case = TRUE)) {
  for (loc in c("C.UTF-8", "en_US.UTF-8", "en_GB.UTF-8")) {
    if (!inherits(try(Sys.setlocale("LC_CTYPE", loc), silent = TRUE), "try-error") &&
        grepl("utf-?8", Sys.getlocale("LC_CTYPE"), ignore.case = TRUE)) break
  }
}
IS_HTML <- knitr::is_html_output()       # plotly only exists in the HTML build

HAS_XGB <- requireNamespace("xgboost", quietly = TRUE)
HAS_GBM <- requireNamespace("gbm",     quietly = TRUE)
if (HAS_XGB) suppressPackageStartupMessages(library(xgboost))
if (!HAS_XGB && HAS_GBM) suppressPackageStartupMessages(library(gbm))
BOOST_ENGINE <- if (HAS_XGB) "XGBoost" else if (HAS_GBM) "GBM" else NA_character_

# ---------------------------------------------------------------------------
# 1. DESIGN SYSTEM
# ---------------------------------------------------------------------------
pal <- list(
  ink       = "#0d2438", ink_soft = "#1b3a55",
  text      = "#10161c", text2    = "#4b5560", muted = "#78838f",
  rule      = "#dcdfda", surface  = "#fcfcfb", surface2 = "#f4f5f3",
  s1 = "#2a78d6", s2 = "#eb6834", s3 = "#1baf7a", s4 = "#eda100", s5 = "#4a3aa7",
  good = "#1baf7a", warn = "#eda100", serious = "#eb6834", critical = "#c0392b",
  seq = c("#cde2fb", "#9ec5f4", "#6da7ec", "#3987e5", "#256abf", "#184f95", "#0d366b")
)
cat_pal <- unname(unlist(pal[c("s1", "s2", "s3", "s4", "s5")]))

theme_clinic <- function(base_size = 12, grid = "y") {
  th <- theme_minimal(base_size = base_size) +
    theme(
      text             = element_text(colour = pal$text),
      plot.title       = element_text(face = "bold", size = base_size * 1.14,
                                      colour = pal$ink, margin = margin(b = 3)),
      plot.subtitle    = element_text(size = base_size * 0.92, colour = pal$text2,
                                      margin = margin(b = 13), lineheight = 1.15),
      plot.caption     = element_text(size = base_size * 0.75, colour = pal$muted,
                                      hjust = 0, margin = margin(t = 12), lineheight = 1.2),
      plot.title.position  = "plot",
      plot.caption.position = "plot",
      axis.title       = element_text(size = base_size * 0.82, colour = pal$text2),
      axis.title.x     = element_text(margin = margin(t = 8)),
      axis.title.y     = element_text(margin = margin(r = 8)),
      axis.text        = element_text(size = base_size * 0.82, colour = pal$text2),
      axis.line.x      = element_line(colour = pal$ink, linewidth = 0.45),
      axis.ticks.x     = element_line(colour = pal$ink, linewidth = 0.35),
      axis.ticks.length = unit(2.5, "pt"),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_line(colour = pal$rule, linewidth = 0.4),
      legend.position  = "top",
      legend.justification = "left",
      legend.title     = element_text(size = base_size * 0.80, colour = pal$text2),
      legend.text      = element_text(size = base_size * 0.80, colour = pal$text2),
      legend.key.height = unit(10, "pt"),
      legend.margin    = margin(b = 2),
      strip.text       = element_text(face = "bold", size = base_size * 0.85,
                                      colour = pal$ink, hjust = 0,
                                      margin = margin(b = 5, t = 3)),
      plot.background  = element_rect(fill = pal$surface, colour = NA),
      panel.background = element_rect(fill = pal$surface, colour = NA),
      plot.margin      = margin(6, 10, 6, 6)
    )
  if (grid == "x") th <- th + theme(panel.grid.major.y = element_blank(),
                                    panel.grid.major.x = element_line(colour = pal$rule, linewidth = 0.4),
                                    axis.line.x = element_blank(), axis.ticks.x = element_blank(),
                                    axis.line.y = element_line(colour = pal$ink, linewidth = 0.45))
  if (grid == "none") th <- th + theme(panel.grid.major.y = element_blank())
  th
}
theme_set(theme_clinic())
update_geom_defaults("col",   list(fill = pal$s1))
update_geom_defaults("line",  list(colour = pal$s1, linewidth = 0.75))
update_geom_defaults("point", list(colour = pal$s1, size = 2.1))
update_geom_defaults("text",  list(family = "", colour = pal$text2, size = 3.0))

# ---------------------------------------------------------------------------
# 2. FORMATTING HELPERS
# ---------------------------------------------------------------------------
gbp   <- function(x, d = 0) paste0("\u00a3", formatC(x, format = "f", big.mark = ",", digits = d))
gbp_k <- function(x)        paste0("\u00a3", formatC(x / 1000, format = "f", digits = 1), "k")
pc    <- function(x, d = 1) paste0(formatC(100 * x, format = "f", digits = d), "%")
pp    <- function(x, d = 1) paste0(formatC(100 * x, format = "f", digits = d), " pp")
num   <- function(x, d = 0) formatC(x, format = "f", big.mark = ",", digits = d)

tbl <- function(x, caption = NULL, ...) {
  k <- knitr::kable(x, caption = caption, booktabs = TRUE, ...)
  if (IS_HTML) {
    k <- kableExtra::kable_styling(
      k, bootstrap_options = c("hover", "condensed"),
      full_width = TRUE, position = "left", font_size = 13.5)
  }
  k
}

show_interactive <- function(p, tooltip = c("x", "y"), legend = FALSE) {
  ttl <- p$labels$title; sub <- p$labels$subtitle; cap <- p$labels$caption
  g <- plotly::ggplotly(p + labs(title = NULL, subtitle = NULL, caption = NULL),
                        tooltip = tooltip)
  plotly::layout(
    g,
    showlegend = legend,
    title = list(
      text = paste0("<b>", ttl, "</b>",
                    if (!is.null(sub)) paste0("<br><sup style='color:#4b5560'>", sub, "</sup>") else ""),
      x = 0, xanchor = "left", font = list(size = 15, color = "#0d2438")),
    margin = list(t = 76, l = 8, r = 8, b = 52),
    hoverlabel = list(bgcolor = "#0d2438", font = list(color = "white", size = 11)),
    plot_bgcolor = "#fcfcfb", paper_bgcolor = "#fcfcfb"
  ) |> plotly::config(displayModeBar = FALSE, responsive = TRUE)
}

wilson <- function(k, n, conf = 0.95) {
  z <- qnorm(1 - (1 - conf) / 2); p <- k / n
  d <- 1 + z^2 / n
  c(lo = (p + z^2 / (2 * n) - z * sqrt(p * (1 - p) / n + z^2 / (4 * n^2))) / d,
    hi = (p + z^2 / (2 * n) + z * sqrt(p * (1 - p) / n + z^2 / (4 * n^2))) / d)
}
rate_ci <- function(dt, by) {
  out <- dt[, .(n = .N, k = sum(no_show)), by = by]
  ci  <- t(mapply(wilson, out$k, out$n))
  out[, `:=`(rate = k / n, lo = ci[, 1], hi = ci[, 2])][]
}
```

```{r}
#| label: ingest
#| include: false

data_path <- "C:/Users/nyabe/OneDrive - american.edu/Desktop/Github/Project_5/uk_private_clinic_no_show_2026.csv"
if (!file.exists(data_path)) data_path <- "uk_private_clinic_no_show_2026.csv"
stopifnot(file.exists(data_path))

raw <- data.table::fread(data_path)

N_APPT    <- nrow(raw)
N_PATIENT <- uniqueN(raw$patient_id)
N_CLINIC  <- uniqueN(raw$clinic_id)
BASE_RATE <- mean(raw$no_show)
```

## Executive summary {.unnumbered}

```{r}
#| label: kpi-headline
#| echo: false

gross_loss  <- sum(raw$estimated_revenue_loss_gbp)
idle_cost   <- sum(raw$staff_idle_cost_gbp)
gross_expo  <- gross_loss + idle_cost
rebook_rate <- mean(raw[no_show == 1]$slot_rebooked_flag)
recovered   <- sum(raw[no_show == 1 & slot_rebooked_flag == 1]$estimated_revenue_loss_gbp)
net_loss    <- gross_expo - recovered
loss_per_appt <- net_loss / N_APPT

kpis <- data.frame(
  label = c("Appointments analysed", "No-show rate", "Gross economic exposure",
            "Slots recovered by rebooking", "Net economic loss", "Net loss per appointment booked"),
  value = c(num(N_APPT), pc(BASE_RATE), gbp(gross_expo),
            pc(rebook_rate), gbp(net_loss), gbp(loss_per_appt, 2)),
  note  = c(paste0(num(N_PATIENT), " patients across ", N_CLINIC, " sites"),
            paste0("1 in every ", formatC(1 / BASE_RATE, digits = 1, format = "f"), " appointments"),
            paste0("fees ", gbp(gross_loss), " plus idle staff ", gbp(idle_cost)),
            paste0(num(sum(raw$slot_rebooked_flag)), " of ", num(sum(raw$no_show)), " missed slots refilled"),
            "gross exposure net of rebooking recovery",
            "the figure that belongs in the operating budget"),
  tone  = c("neutral", "critical", "serious", "good", "critical", "serious"),
  stringsAsFactors = FALSE
)

# A three-column layout with one very short numeric column squeezed between two
# long text columns causes Word's table auto-fit to wrap the number itself
# mid-digit, and it is simplest to use one consistent table in both formats
# rather than a custom HTML tile layout that depends on the stylesheet loading.
kpis_tbl <- data.frame(Measure = kpis$label,
                       Detail = paste0("**", kpis$value, "** : ", kpis$note))
tbl(kpis_tbl, caption = "Exhibit 1. Headline measures, FY2026 booking year",
    col.names = c("Measure", "Value and basis"))
```

This report answers one question for the clinic network's leadership: **why are appointments being
missed, and what is it worth to fix that?** A five-site network booked `r num(N_APPT)` appointments in
the year covered by this extract, and `r pc(BASE_RATE)` of them, one in every
`r formatC(1 / BASE_RATE, digits = 1, format = "f")`, went unattended. After crediting back the slots
that were successfully rebooked, that leaves `r gbp(net_loss)` of real economic loss a year, or
`r gbp(loss_per_appt, 2)` on every single appointment the network takes.

No-show risk here is driven by three things the clinic already controls: **how far ahead an
appointment is booked**, **how the reminder programme is run**, and **whether a cancelled slot is
released back into the booking system**. It is not driven by who the patient is: age, gender, clinic
site and day of week make essentially no difference. A predictive model built from the operational
factors ranks appointments by risk well enough to make a targeted call programme profitable, and a
costed four-part action plan below recovers a substantial share of the loss, most of it starting next
quarter with no pilot required.

> **How to read this report.** It has been deliberately shortened to ten pages for a general
> business audience. Every figure quoted is calculated directly from the underlying data by the code
> that also produced the reader's copy of this document, not typed in by hand, so every number here
> can be traced back to its source. A companion technical report covers the full statistical detail,
> including 41 supporting exhibits, for readers who want it.

# The data behind this report

Before analysing anything, the ten-thousand-row booking file was checked for problems, because a
model is only as trustworthy as what it is built on. Three issues were found and corrected. Two
columns that were supposed to record each patient's history of missed appointments turn out to be
completely empty (`r uniqueN(raw$patient_previous_noshow_count)` distinct value across all
`r num(N_APPT)` rows), so this analysis cannot flag "a patient who has a habit of missing
appointments", only "an appointment that looks risky right now," which turns out to matter more in
practice. Three financial columns (money lost, staff time wasted, whether the slot was rebooked) are
calculated directly from whether the appointment was missed; using them to *predict* a no-show would
be circular reasoning, so they are used only to size the loss in pounds, never to feed the model.
Finally, the reminder-tracking fields overlapped in a way that would have broken the statistics if
left as supplied, so they were consolidated into one clean signal. None of this changes the
conclusions; it is what makes the conclusions trustworthy.

```{r}
#| label: prep
#| include: false

d <- copy(raw)
d[, c("patient_previous_noshow_count", "patient_noshow_rate") := NULL]

d[, `:=`(
  reminder_response = factor(reminder_response,
                             levels = c("Confirmed", "Ignored", "Cancelled", "No Reminder")),
  reminder_lead     = factor(reminder_hours_before, levels = c(24, 48, 72),
                             labels = c("24h", "48h", "72h")),
  clinic_type       = factor(clinic_type, levels = c("GP", "Dental", "Physiotherapy", "Dermatology")),
  appointment_type  = factor(appointment_type, levels = c("Follow-up", "New Patient")),
  gender            = factor(gender, levels = c("Female", "Male")),
  clinic_id         = factor(clinic_id),
  dow               = factor(appointment_day_of_week,
                             levels = c("Monday", "Tuesday", "Wednesday", "Thursday",
                                        "Friday", "Saturday", "Sunday")),
  outcome           = factor(no_show, levels = c(0, 1), labels = c("Attended", "No-show"))
)]

d[, `:=`(
  lead_band = cut(days_in_advance, breaks = c(-1, 3, 7, 14, 21, Inf),
                  labels = c("0-3 d", "4-7 d", "8-14 d", "15-21 d", "22+ d")),
  dist_band = cut(distance_to_clinic_km, breaks = c(-1, 2, 5, 10, Inf),
                  labels = c("0-2 km", "2-5 km", "5-10 km", "10+ km")),
  age_band  = cut(age, breaks = c(17, 29, 39, 49, 59, Inf),
                  labels = c("18-29", "30-39", "40-49", "50-59", "60+")),
  gap_band  = cut(days_since_last_visit, breaks = c(-1, 60, 120, 240, Inf),
                  labels = c("<2 mo", "2-4 mo", "4-8 mo", "8+ mo"))
)]

d[, `:=`(
  rem_48h    = as.integer(sms_reminder_sent == 1 & reminder_hours_before == 48),
  rem_72h    = as.integer(sms_reminder_sent == 1 & reminder_hours_before == 72),
  rem_second = as.integer(sms_reminder_sent == 1 & number_of_reminders_sent == 2)
)]

d[, `:=`(
  gross_exposure = estimated_revenue_loss_gbp + staff_idle_cost_gbp,
  net_loss       = fifelse(slot_rebooked_flag == 1, staff_idle_cost_gbp,
                           estimated_revenue_loss_gbp + staff_idle_cost_gbp)
)]
```

# Sizing the loss

```{r}
#| label: fig-waterfall
#| echo: false
#| fig-cap: "Exhibit 2. From gross fee exposure to net economic loss. Rebooking recovers a meaningful share of forgone fees, but the idle-staff charge is incurred whether or not the slot is refilled."
#| fig-height: 4.6

wf <- data.table(
  step = factor(c("Forgone\nfees", "Idle clinician\nand room", "Recovered by\nrebooking", "Net economic\nloss"),
                levels = c("Forgone\nfees", "Idle clinician\nand room", "Recovered by\nrebooking", "Net economic\nloss")),
  value = c(gross_loss, idle_cost, -recovered, net_loss),
  kind  = c("add", "add", "subtract", "total")
)
wf[, end := cumsum(fifelse(kind == "total", 0, value))]
wf[kind == "total", end := net_loss]
wf[, start := fifelse(kind == "total", 0, end - value)]
wf[, xi := as.numeric(step)]

ggplot(wf) +
  geom_rect(aes(xmin = xi - 0.34, xmax = xi + 0.34, ymin = start, ymax = end, fill = kind),
            colour = pal$surface, linewidth = 0.9) +
  geom_segment(data = wf[1:3], aes(x = xi + 0.34, xend = xi + 0.66, y = end, yend = end),
               colour = pal$muted, linewidth = 0.35, linetype = "22") +
  geom_text(aes(x = xi, y = pmax(start, end),
                label = paste0(ifelse(value < 0, "-", ""), gbp(abs(value)))),
            vjust = -0.75, family = "", size = 3.35, fontface = "bold", colour = pal$ink) +
  scale_fill_manual(values = c(add = pal$s1, subtract = pal$s3, total = pal$critical), guide = "none") +
  scale_x_continuous(breaks = 1:4, labels = levels(wf$step)) +
  scale_y_continuous(labels = function(x) gbp_k(x), expand = expansion(mult = c(0, 0.17))) +
  labs(title = "Only part of the headline exposure is genuinely lost",
       subtitle = paste0("Gross exposure of ", gbp(gross_expo), " nets down to ", gbp(net_loss),
                         " once the ", pc(rebook_rate), " of missed slots that get refilled are credited back"),
       x = NULL, y = "Value, FY2026",
       caption = paste0("n = ", num(N_APPT), " appointments, of which ", num(sum(d$no_show)),
                        " were no-shows. Idle cost is fixed at 35% of the forgone fee.")) +
  theme_clinic()
```

The headline figure of `r gbp(gross_loss)` in lost fees is the wrong number to plan around, in two
directions at once. It understates the true cost because every missed appointment also strands a
clinician and a room, adding `r gbp(idle_cost)` in idle time. It overstates the true cost because
`r pc(rebook_rate)` of missed slots get filled by another patient, so that fee is not actually lost.
Netting both adjustments gives `r gbp(net_loss)`, or `r gbp(loss_per_appt, 2)` per appointment booked.
That per-appointment figure is the one that matters for planning, because it is the one that scales
correctly if the network grows: it stays true whether the network books ten thousand appointments a
year or twenty thousand, while the lump-sum total does not.

Loss is not spread evenly. The *rate* of no-shows barely differs across the four service lines, but
the *cost* of a no-show does, because a missed dermatology appointment carries a higher fee than a
missed GP appointment. A prevention budget spent in proportion to failure rate would therefore waste
money; spent in proportion to pounds at risk, it goes where it is worth most. That is how the
targeting model described below is scored.

# What drives it

```{r}
#| label: drivers-calc
#| include: false

rt <- rate_ci(d[sms_reminder_sent == 1], "reminder_lead")
rt[, reminder_lead := factor(reminder_lead, levels = c("24h", "48h", "72h"))]

pt_72 <- prop.test(
  x = c(sum(d[sms_reminder_sent == 1 & reminder_hours_before == 72]$no_show),
        sum(d[sms_reminder_sent == 1 & reminder_hours_before < 72]$no_show)),
  n = c(nrow(d[sms_reminder_sent == 1 & reminder_hours_before == 72]),
        nrow(d[sms_reminder_sent == 1 & reminder_hours_before < 72])),
  correct = FALSE)
eff_72 <- diff(rev(pt_72$estimate))

ch <- rate_ci(d, "has_chronic_condition_flag")
ch[, lab := factor(has_chronic_condition_flag, levels = c(0, 1),
                   labels = c("No chronic condition", "Chronic condition recorded"))]
chi_ch <- chisq.test(table(d$has_chronic_condition_flag, d$no_show))
or_ch  <- { tt <- table(d$has_chronic_condition_flag, d$no_show)
            (tt[2, 2] * tt[1, 1]) / (tt[2, 1] * tt[1, 2]) }

t2b <- prop.test(x = c(sum(d[number_of_reminders_sent == 2]$no_show),
                        sum(d[number_of_reminders_sent == 1]$no_show)),
                  n = c(nrow(d[number_of_reminders_sent == 2]), nrow(d[number_of_reminders_sent == 1])),
                  correct = FALSE)
```

```{r}
#| label: fig-reminder-response
#| echo: false
#| fig-cap: "Exhibit 3. No-show rate by how the patient responded to their reminder. The gap is large, but because patients choose whether to confirm, it is an upper bound on what a confirmation campaign could achieve, not a clean effect estimate."
#| fig-height: 4.2

rr <- rate_ci(d, "reminder_response")
rr[, reminder_response := factor(reminder_response, levels = rr[order(rate)]$reminder_response)]
chi_rr <- chisq.test(table(d$reminder_response, d$no_show))

ggplot(rr, aes(x = reminder_response, y = rate)) +
  geom_hline(yintercept = BASE_RATE, colour = pal$muted, linewidth = 0.4, linetype = "22") +
  geom_col(aes(fill = rate), width = 0.62) +
  annotate("text", x = 0.56, y = BASE_RATE, label = paste0("network ", pc(BASE_RATE)),
           vjust = -0.55, hjust = 0, family = "", size = 2.85, colour = pal$muted) +
  geom_linerange(aes(ymin = lo, ymax = hi), colour = pal$ink, linewidth = 0.65, alpha = 0.55) +
  geom_text(aes(y = hi, label = pc(rate)), vjust = -1.30, family = "", size = 3.2,
            fontface = "bold", colour = pal$ink) +
  scale_fill_gradientn(colours = pal$seq[c(2, 4, 6)], guide = "none") +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.345),
                     expand = expansion(mult = c(0, 0.03))) +
  labs(title = "Confirmation separates the book almost in half",
       subtitle = paste0("Confirmers fail at ", pc(rr[reminder_response == "Confirmed"]$rate),
                         " against ", pc(rr[reminder_response == "Cancelled"]$rate), " for cancellers left on the book"),
       x = NULL, y = "No-show rate",
       caption = "Vertical lines are 95% confidence intervals.") +
  theme_clinic()
```

Three operational factors carry almost all of the explanatory power. **Booking lead time**:
`r pc(mean(d[days_in_advance >= 22]$no_show))` of appointments made 22 or more days out are missed,
against `r pc(mean(d[days_in_advance <= 3]$no_show))` for those booked within three days, and risk
rises step by step in between. **Reminder confirmation**, shown above: patients who
confirm miss `r pc(rr[reminder_response == "Confirmed"]$rate)` of appointments; patients who cancel
but are left on the book anyway miss `r pc(rr[reminder_response == "Cancelled"]$rate)`. That gap is
real but partly reflects who bothers to confirm, not just what confirming does, so it is used here
for targeting rather than as a promised effect. **Reminder timing** is the cleaner result: because
reminders are sent at 24, 48 or 72 hours in roughly equal numbers regardless of patient type, a
72-hour reminder is associated with a `r pp(abs(eff_72))` lower no-show rate than a 24 or 48-hour one
(p `r ifelse(pt_72$p.value < 0.0001, "< 0.0001", paste0("= ", formatC(pt_72$p.value, digits = 4, format = "f")))`),
and this one comes closest to a genuine cause-and-effect finding, since the clinic assigns the timing,
not the patient.

Two findings are worth flagging precisely because they run against instinct. Patients with a chronic
condition on file are *more* reliable attenders, not less
(`r pc(ch[has_chronic_condition_flag == 1]$rate)` versus
`r pc(ch[has_chronic_condition_flag == 0]$rate)`, odds ratio
`r formatC(or_ch, digits = 2, format = "f")`), because an ongoing treatment relationship gives them a
reason to show up. And several plausible levers turn out not to matter at all: a second reminder
message changes the no-show rate by only
`r pp(abs(t2b$estimate[1] - t2b$estimate[2]))`, statistically indistinguishable from zero, and neither
clinic site nor day of week nor patient age moves risk in any usable way. Reporting a null result is
as valuable as reporting a real one, because it says where *not* to spend.

# Building the model

A single prediction is not enough, because a clinic can act on risk at two different moments with two
different sets of information: at the moment the appointment is booked, and again in the 24 to 72
hours before it, once the reminder response is known. Two versions of the model were built to match
those two moments.

```{r}
#| label: split
#| include: false

pids  <- unique(d$patient_id)
perm  <- sample(pids)
n_te  <- round(0.25 * length(perm)); n_ca <- round(0.25 * length(perm))
assign_split <- data.table(
  patient_id = perm,
  split = c(rep("test", n_te), rep("calibration", n_ca),
            rep("train", length(perm) - n_te - n_ca)))
d[assign_split, split := i.split, on = "patient_id"]
d[, split := factor(split, levels = c("train", "calibration", "test"))]

train <- d[split == "train"]
calib <- d[split == "calibration"]
test  <- d[split == "test"]

booking_vars <- c("age", "gender", "distance_to_clinic_km", "clinic_type", "clinic_id",
                  "appointment_type", "consultation_fee_gbp", "days_in_advance",
                  "appointment_hour", "is_weekend", "is_peak_hour", "dow",
                  "patient_total_appointments", "days_since_last_visit",
                  "has_chronic_condition_flag")
pre_vars <- c(booking_vars, "reminder_response", "rem_48h", "rem_72h", "rem_second")

design <- function(vars, data) {
  f <- stats::as.formula(paste("~", paste(vars, collapse = " + ")))
  stats::model.matrix(f, data)[, -1, drop = FALSE]
}

K <- 10
fold_map  <- data.table(patient_id = unique(train$patient_id))
fold_map[, fold := sample(rep_len(1:K, .N))]
train_fold <- fold_map[match(train$patient_id, fold_map$patient_id)]$fold
```

Every patient's appointments were kept together in only one of three groups: a training group to fit
the models, a calibration group to check and correct the models' confidence, and a holdout group set
aside untouched until the very last step, so that the final accuracy figure is a genuinely honest
one. Four competing approaches were tried, from a plain, fully transparent regression up to a tuned
machine-learning ensemble, each tuned by cross-validation (repeatedly testing settings against a slice
of the training data itself, never against the holdout). The model that is actually deployed was
chosen on the calibration group, using a formal statistical test: find the best-scoring candidate,
keep every other candidate that cannot be shown to be meaningfully worse, and among those, deploy the
simplest and most explainable one. That rule exists to stop the choice of model from depending on
which one happened to get lucky on a single sample.

```{r}
#| label: modelling
#| include: false

fit_stage <- function(vars, stage_name) {

  Xtr <- design(vars, train); Xte <- design(vars, test); Xca <- design(vars, calib)
  ytr <- train$no_show;       yte <- test$no_show;        yca <- calib$no_show
  assign_map <- attr(stats::model.matrix(
    stats::as.formula(paste("~", paste(vars, collapse = " + "))), train), "assign")[-1]

  ## --- M1: unpenalised logistic ------------------------------------------
  m_lr  <- stats::glm(y ~ ., data = data.frame(y = ytr, Xtr), family = stats::binomial)
  p_lr  <- stats::predict(m_lr, newdata = data.frame(Xte), type = "response")
  c_lr  <- stats::predict(m_lr, newdata = data.frame(Xca), type = "response")

  ## --- M2: elastic net, grouped 10-fold CV over alpha and lambda ----------
  alphas <- c(0, 0.25, 0.5, 0.75, 1)
  en_grid <- data.table(alpha = alphas, cv_auc = NA_real_, lambda_1se = NA_real_, nonzero = NA_integer_)
  en_fits <- vector("list", length(alphas))
  for (i in seq_along(alphas)) {
    cvf <- glmnet::cv.glmnet(Xtr, ytr, family = "binomial", alpha = alphas[i],
                             foldid = train_fold, type.measure = "auc", standardize = TRUE)
    en_fits[[i]] <- cvf
    j <- which(cvf$lambda == cvf$lambda.1se)
    en_grid[i, `:=`(cv_auc = cvf$cvm[j], lambda_1se = cvf$lambda.1se, nonzero = cvf$nzero[j])]
  }
  bi   <- which.max(en_grid$cv_auc)
  m_en <- en_fits[[bi]]
  p_en <- as.numeric(stats::predict(m_en, Xte, s = "lambda.1se", type = "response"))
  c_en <- as.numeric(stats::predict(m_en, Xca, s = "lambda.1se", type = "response"))

  ## --- M3: random forest, grid search on out-of-bag AUC ------------------
  rf_grid <- data.table(expand.grid(mtry = c(3, 6, 10),
                                    min.node.size = c(50, 150, 400, 900, 1600)))
  rf_grid[, oob_auc := NA_real_]
  for (i in seq_len(nrow(rf_grid))) {
    f <- ranger::ranger(x = Xtr, y = factor(ytr), num.trees = 500,
                        mtry = rf_grid$mtry[i], min.node.size = rf_grid$min.node.size[i],
                        probability = TRUE, seed = 7, num.threads = 2)
    rf_grid[i, oob_auc := as.numeric(pROC::auc(pROC::roc(ytr, f$predictions[, 2], quiet = TRUE)))]
  }
  ri   <- which.max(rf_grid$oob_auc)
  m_rf <- ranger::ranger(x = Xtr, y = factor(ytr), num.trees = 1500,
                         mtry = rf_grid$mtry[ri], min.node.size = rf_grid$min.node.size[ri],
                         probability = TRUE, importance = "permutation", seed = 7, num.threads = 2)
  p_rf <- stats::predict(m_rf, Xte)$predictions[, 2]
  c_rf <- stats::predict(m_rf, Xca)$predictions[, 2]

  ## --- M4: gradient boosting --------------------------------------------
  boost_note   <- NA_character_
  boost_engine <- NA_character_
  p_bo <- c_bo <- NULL; imp_bo <- NULL

  if (HAS_XGB) {
    ok <- try({
      fold_list <- split(seq_len(nrow(Xtr)), train_fold)
      prm <- list(objective = "binary:logistic", eval_metric = "auc",
                  eta = 0.03, max_depth = 3, subsample = 0.8,
                  colsample_bytree = 0.8, min_child_weight = 30)
      dtr <- xgboost::xgb.DMatrix(data = Xtr, label = ytr)
      cvb <- xgboost::xgb.cv(params = prm, data = dtr, nrounds = 1500, folds = fold_list,
                             early_stopping_rounds = 60, verbose = 0)
      best_n <- which.max(cvb$evaluation_log$test_auc_mean)
      m_bo   <- xgboost::xgb.train(params = prm, data = dtr, nrounds = best_n, verbose = 0)
      p_bo   <- as.numeric(stats::predict(m_bo, xgboost::xgb.DMatrix(data = Xte)))
      c_bo   <- as.numeric(stats::predict(m_bo, xgboost::xgb.DMatrix(data = Xca)))
      imp_bo <- as.data.table(xgboost::xgb.importance(model = m_bo))[, .(Feature, Gain)]
      boost_engine <- "XGBoost"
      boost_note <- paste0("eta 0.03, depth 3, subsample 0.8, column sample 0.8, minimum child ",
                           "weight 30, and ", best_n,
                           " trees selected by grouped 10-fold cross-validation with early stopping")
      TRUE
    }, silent = TRUE)
    if (inherits(ok, "try-error")) { p_bo <- NULL; boost_engine <- NA_character_ }
  }

  if (is.null(p_bo) && HAS_GBM) {
    dfb  <- data.frame(y = ytr, Xtr)
    m_bo <- gbm::gbm(y ~ ., data = dfb, distribution = "bernoulli", n.trees = 1500,
                     interaction.depth = 3, shrinkage = 0.03, bag.fraction = 0.8,
                     n.minobsinnode = 30, cv.folds = 5, verbose = FALSE, n.cores = 2)
    best_n <- gbm::gbm.perf(m_bo, method = "cv", plot.it = FALSE)
    p_bo   <- stats::predict(m_bo, newdata = data.frame(Xte), n.trees = best_n, type = "response")
    c_bo   <- stats::predict(m_bo, newdata = data.frame(Xca), n.trees = best_n, type = "response")
    ib     <- as.data.table(summary(m_bo, n.trees = best_n, plotit = FALSE))
    imp_bo <- ib[, .(Feature = as.character(var), Gain = rel.inf / 100)]
    boost_engine <- "GBM"
    boost_note <- paste0("shrinkage 0.03, depth 3, bagging fraction 0.8, minimum node 30, and ",
                         best_n, " trees selected by 5-fold cross-validation")
  }

  if (is.null(p_bo)) { p_bo <- rep(mean(ytr), length(yte)); c_bo <- rep(mean(ytr), length(yca)) }

  preds <- list(`Logistic regression` = p_lr, `Elastic net` = p_en, `Random forest` = p_rf)
  cals  <- list(`Logistic regression` = c_lr, `Elastic net` = c_en, `Random forest` = c_rf)

  pfun <- list(
    `Logistic regression` = function(M) as.numeric(stats::predict(m_lr, newdata = data.frame(M),
                                                                  type = "response")),
    `Elastic net`         = function(M) as.numeric(stats::predict(m_en, M, s = "lambda.1se",
                                                                  type = "response")),
    `Random forest`       = function(M) stats::predict(m_rf, M)$predictions[, 2])

  if (!is.na(boost_engine)) {
    bnm <- paste0("Gradient boosting (", boost_engine, ")")
    preds[[bnm]] <- p_bo
    cals[[bnm]]  <- c_bo
    pfun[[bnm]]  <- if (boost_engine == "XGBoost") {
      function(M) as.numeric(stats::predict(m_bo, xgboost::xgb.DMatrix(data = M)))
    } else {
      function(M) as.numeric(stats::predict(m_bo, newdata = data.frame(M),
                                            n.trees = best_n, type = "response"))
    }
  }

  list(stage = stage_name, vars = vars, X_test = Xte, y_test = yte, assign_map = assign_map,
       preds = preds, cals = cals, pfun = pfun, y_calib = yca, boost_engine = boost_engine,
       m_lr = m_lr, m_en = m_en, m_rf = m_rf,
       en_grid = en_grid, rf_grid = rf_grid, en_alpha = alphas[bi],
       rf_mtry = rf_grid$mtry[ri], rf_node = rf_grid$min.node.size[ri],
       boost_note = boost_note, imp_bo = imp_bo)
}

evaluate <- function(y, p) {
  r  <- pROC::roc(y, p, quiet = TRUE, direction = "<")
  ci <- as.numeric(pROC::ci.auc(r))
  lp  <- stats::qlogis(pmin(pmax(p, 1e-6), 1 - 1e-6))
  cal <- stats::glm(y ~ lp, family = stats::binomial)
  cs  <- summary(cal)$coefficients
  data.table(AUC = as.numeric(pROC::auc(r)), AUC_lo = ci[1], AUC_hi = ci[3],
             Brier = mean((p - y)^2),
             Cal_slope = cs[2, 1],
             Cal_slope_lo = cs[2, 1] - 1.96 * cs[2, 2],
             Cal_slope_hi = cs[2, 1] + 1.96 * cs[2, 2],
             Cal_int   = cs[1, 1],
             KS = max(abs(stats::ecdf(p[y == 1])(sort(p)) - stats::ecdf(p[y == 0])(sort(p)))))
}

stageA <- fit_stage(booking_vars, "Stage A: booking-time")
stageB <- fit_stage(pre_vars,     "Stage B: pre-appointment")
BOOST_ENGINE <- stageB$boost_engine

perf <- rbindlist(lapply(list(stageA, stageB), function(s)
  rbindlist(lapply(names(s$preds), function(nm)
    cbind(data.table(Stage = s$stage, Model = nm), evaluate(s$y_test, s$preds[[nm]]))))))

simplicity <- function(m) fcase(grepl("^Logistic",  m), 1L,
                                grepl("^Elastic",   m), 2L,
                                grepl("forest",     m), 3L,
                                default = 4L)

roc_cal <- lapply(stageB$cals, function(p)
  pROC::roc(stageB$y_calib, p, quiet = TRUE, direction = "<"))
sel_dt <- data.table(Model = names(roc_cal),
                     cal_auc = vapply(roc_cal, function(r) as.numeric(pROC::auc(r)), numeric(1)))
setorder(sel_dt, -cal_auc)
TOP_MODEL <- sel_dt$Model[1]
sel_dt[, delong_p := vapply(Model, function(m) if (m == TOP_MODEL) NA_real_ else
      pROC::roc.test(roc_cal[[TOP_MODEL]], roc_cal[[m]], method = "delong")$p.value, numeric(1))]
sel_dt[, tied   := is.na(delong_p) | delong_p > 0.05]
sel_dt[, simple := simplicity(Model)]

SEL_MODEL  <- sel_dt[tied == TRUE][order(simple, -cal_auc)]$Model[1]
N_RIVALS   <- nrow(sel_dt) - 1L
N_TIED     <- sum(sel_dt$tied) - 1L
best_row   <- perf[Stage == "Stage B: pre-appointment" & Model == SEL_MODEL]
BEST_MODEL <- SEL_MODEL
raw_p      <- stageB$preds[[SEL_MODEL]]
auc_gain   <- perf[Model == SEL_MODEL & Stage == "Stage B: pre-appointment"]$AUC -
              perf[Model == SEL_MODEL & Stage == "Stage A: booking-time"]$AUC
SEL_IS_LINEAR <- simplicity(SEL_MODEL) <= 2L

clip     <- function(p) pmin(pmax(p, 1e-6), 1 - 1e-6)
platt    <- stats::glm(y ~ lo, family = stats::binomial,
                       data = data.frame(y = stageB$y_calib,
                                         lo = stats::qlogis(clip(stageB$cals[[BEST_MODEL]]))))
recalibrate <- function(p) as.numeric(stats::predict(
  platt, newdata = data.frame(lo = stats::qlogis(clip(p))), type = "response"))
best_p   <- recalibrate(raw_p)

cal_before <- evaluate(stageB$y_test, raw_p)
cal_after  <- evaluate(stageB$y_test, best_p)
```

```{r}
#| label: exhibit-performance-short
#| echo: false

show2 <- perf[Stage == "Stage B: pre-appointment"]
show2 <- merge(show2, sel_dt[, .(Model, tied)], by = "Model", all.x = TRUE, sort = FALSE)
setorder(show2, -AUC)
show2 <- show2[, .(Model,
                   `Holdout accuracy score (AUC)` = formatC(AUC, digits = 3, format = "f"),
                   `95% confidence range` = paste0(formatC(AUC_lo, digits = 3, format = "f"), " to ",
                                                   formatC(AUC_hi, digits = 3, format = "f")),
                   Verdict = fifelse(Model == SEL_MODEL, "Deployed",
                              fifelse(tied, "Statistically tied", "Measurably weaker")))]
tbl(show2, caption = paste0("Exhibit 4. How the four candidate models performed on ", num(nrow(test)),
                            " appointments none of them had seen before. ", SEL_MODEL,
                            " was deployed because it is not measurably worse than the top scorer and is ",
                            "the simplest one a person can read and check by hand."))
```

The score in that table, AUC, works like this: pick one appointment at random that turned out to be a
no-show and one that did not, and ask the model which it thought was riskier. A score of 50% is a coin
flip; a score of 100% would mean the model is never wrong. `r SEL_MODEL` scores
`r formatC(best_row$AUC, digits = 3, format = "f")`, which is honest but modest, and it is reported on
appointments and patients the model never saw during fitting, so it is not a flattered number. The
notable result is that a far more sophisticated machine-learning model, tested alongside it, could not
score measurably better: `r N_TIED` of its `r N_RIVALS` rivals are statistically indistinguishable
from the winner. That is not a disappointing outcome, it is a finding in its own right — it means the
relationship between these operational factors and attendance is simple and additive rather than
hidden in complex combinations, so there is no accuracy paid up in exchange for using the version a
clinician can actually read. Before its probabilities were used for anything financial, the model's
confidence was also double-checked and corrected on data set aside for exactly that purpose, so that a
predicted 25% risk really does behave like roughly a one-in-four chance.

```{r}
#| label: fig-permutation
#| echo: false
#| fig-cap: "Exhibit 5. What actually carries the model. Each factor is scrambled in turn and the resulting drop in accuracy measured on data the model never trained on; a bigger drop means the model relies on that factor more."
#| fig-height: 4.6

perm_importance <- function(stage_obj, pred_fun, n_rep = 20) {
  X <- stage_obj$X_test; y <- stage_obj$y_test
  base_auc <- as.numeric(pROC::auc(pROC::roc(y, pred_fun(X), quiet = TRUE, direction = "<")))
  grp <- split(seq_len(ncol(X)), stage_obj$assign_map)
  nm  <- stage_obj$vars[as.integer(names(grp))]
  out <- vapply(seq_along(grp), function(g) {
    drops <- vapply(seq_len(n_rep), function(r) {
      Xp <- X
      idx <- sample.int(nrow(X))
      Xp[, grp[[g]]] <- X[idx, grp[[g]], drop = FALSE]
      base_auc - as.numeric(pROC::auc(pROC::roc(y, pred_fun(Xp), quiet = TRUE, direction = "<")))
    }, numeric(1))
    c(mean(drops), stats::sd(drops))
  }, numeric(2))
  data.table(variable = nm, drop = out[1, ], se = out[2, ] / sqrt(n_rep))
}

pf_sel <- stageB$pfun[[SEL_MODEL]]
pi_dt  <- perm_importance(stageB, pf_sel, n_rep = 20)

pretty_var <- c(reminder_response = "Reminder response", days_in_advance = "Booking lead time",
                distance_to_clinic_km = "Distance to clinic",
                has_chronic_condition_flag = "Chronic condition flag",
                rem_48h = "Reminder sent 48h ahead", rem_72h = "Reminder sent 72h ahead",
                rem_second = "Second reminder sent", age = "Age",
                consultation_fee_gbp = "Consultation fee", clinic_type = "Service line",
                appointment_hour = "Appointment hour", days_since_last_visit = "Days since last visit",
                patient_total_appointments = "Patient appointment count", gender = "Gender",
                appointment_type = "Appointment type", dow = "Day of week",
                is_weekend = "Weekend flag", is_peak_hour = "Peak-hour flag",
                clinic_id = "Clinic site")
pi_dt[, label := ifelse(variable %in% names(pretty_var), pretty_var[variable], variable)]
setorder(pi_dt, drop)
pi_dt[, label := factor(label, levels = label)]
pi_dt[, tier := fifelse(drop > 0.008, "Carries the model",
                 fifelse(drop > 0.0005, "Marginal contribution", "No measurable contribution"))]
N_ZERO  <- pi_dt[tier == "No measurable contribution", .N]
CARRY   <- rev(as.character(pi_dt[tier != "No measurable contribution"]$label))
N_CARRY <- length(CARRY)
carry_list <- if (N_CARRY > 1)
  paste0(paste(CARRY[-N_CARRY], collapse = ", "), " and ", CARRY[N_CARRY]) else CARRY[1]

ggplot(pi_dt, aes(x = drop, y = label, fill = tier)) +
  geom_col(width = 0.66) +
  geom_linerange(aes(xmin = drop - 1.96 * se, xmax = drop + 1.96 * se),
                 colour = pal$ink, linewidth = 0.45, alpha = 0.6) +
  scale_fill_manual(values = c(`Carries the model` = pal$s1,
                               `Marginal contribution` = pal$seq[3],
                               `No measurable contribution` = pal$rule),
                    name = NULL, breaks = c("Carries the model", "Marginal contribution",
                                            "No measurable contribution")) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1)),
                     labels = number_format(accuracy = 0.005)) +
  labs(title = paste0(N_CARRY, " factors carry the model; the rest contribute nothing measurable"),
       subtitle = paste0("Loss in accuracy when each factor is scrambled, mean of 20 repeats. Model: ", SEL_MODEL),
       x = "Drop in holdout accuracy when scrambled", y = NULL,
       caption = "Related dummy columns from one original factor are scrambled together.") +
  theme_clinic(grid = "x", base_size = 11)
```

Only `r N_CARRY` of the `r nrow(pi_dt)` factors available to the model do any measurable work:
**`r carry_list`**. That shortlist is what makes the problem tractable for a clinic to act on. Lead
time and reminder response are set by the clinic, not by the patient; distance is known the moment a
patient registers and costs nothing to look up. The model is, in effect, telling the network that its
no-show problem is a scheduling and communication problem, not a patient problem.

# The business case

Not every classification error costs the same. A wasted outreach call is cheap; a missed appointment
that goes uncontacted is expensive.

```{r}
#| label: exhibit-cost-matrix
#| include: false

C_OUTREACH  <- 4.00
UPLIFT      <- 0.20
mean_fee    <- mean(d[no_show == 1]$estimated_revenue_loss_gbp)
mean_net    <- mean(d[no_show == 1]$net_loss)
```

On these figures, a genuine no-show costs the clinic about `r gbp(mean_net, 0)` in lost value once idle time and
rebooking are accounted for, while a phone call to a patient who was always going to attend costs
about `r gbp(C_OUTREACH, 2)`. Because those two numbers are roughly
`r formatC(mean_net / C_OUTREACH, digits = 0, format = "f")` to one apart, the sensible policy is not
to call every patient, or the top half, or the top 5%; it is to work out, from the economics, exactly
which risk score makes a call worth its cost, and stop there.

```{r}
#| label: fig-threshold
#| include: false

econ <- data.table(p = best_p, y = stageB$y_test,
                   value = test$estimated_revenue_loss_gbp,
                   fee = test$consultation_fee_gbp)
econ[, net_if_missed := fee * (1 - rebook_rate) + fee * 0.35]

thr_grid <- seq(0.05, 0.60, by = 0.005)
tres <- rbindlist(lapply(thr_grid, function(t) {
  flag <- econ$p >= t
  benefit <- sum(UPLIFT * econ$net_if_missed[flag & econ$y == 1])
  cost    <- C_OUTREACH * sum(flag)
  data.table(threshold = t, flagged = sum(flag), share = mean(flag),
             tp = sum(flag & econ$y == 1), fp = sum(flag & econ$y == 0),
             benefit = benefit, cost = cost, net = benefit - cost)
}))
best_thr <- tres[which.max(net)]
scale_up <- N_APPT / nrow(test)
```

```{r}
#| label: exhibit-confusion
#| include: false

flag <- econ$p >= best_thr$threshold
cmx  <- table(factor(flag, c(TRUE, FALSE)), factor(econ$y, c(1, 0)))
sens <- cmx[1, 1] / sum(cmx[, 1]); spec <- cmx[2, 2] / sum(cmx[, 2])
prec <- cmx[1, 1] / sum(cmx[1, ])
```

At the economically optimal cut-off, `r pc(best_thr$threshold, 1)` predicted risk, the clinic would
contact `r pc(best_thr$share)` of its book and catch `r pc(sens)` of the appointments that go on to be
missed, for a modelled net benefit of `r gbp(best_thr$net * scale_up)` a year. Only about
`r pc(prec)` of the calls made at that threshold turn out to be genuine no-shows, which sounds low but
is not a flaw: because the value of catching a real no-show is so much larger than the cost of a
wasted call, the programme is still comfortably profitable even with most calls "missing." Two of the
numbers behind that estimate, the cost of a call and how often a call actually changes a patient's
behaviour, are assumptions rather than measurements, because this dataset contains no record of anyone
having actually tried the intervention. The chart below shows what happens to the business case across
a wide range of both assumptions, rather than hiding that uncertainty behind a single confident figure.

```{r}
#| label: fig-sensitivity
#| echo: false
#| fig-cap: "Exhibit 6. Modelled annual net benefit of the outreach programme, re-optimised for every combination of call cost and behavioural effect. A cell reading \"do not run\" means no threshold covers its costs, so the clinic is better off not calling anyone under that assumption."
#| fig-height: 4.8

grid_u <- seq(0.05, 0.40, by = 0.05)
grid_c <- c(2, 4, 6, 8, 10, 12)
sens_dt <- rbindlist(lapply(grid_u, function(u) rbindlist(lapply(grid_c, function(cc) {
  vals <- vapply(thr_grid, function(t) {
    fl <- econ$p >= t
    u * sum(econ$net_if_missed[fl & econ$y == 1]) - cc * sum(fl)
  }, numeric(1))
  best <- max(max(vals), 0)
  data.table(uplift = u, cost = cc, net = best * scale_up)
}))))
sens_dt[, worth_running := net > 500]
BASE_CELL  <- sens_dt[uplift == UPLIFT & cost == C_OUTREACH]$net
SHARE_LIVE <- mean(sens_dt$worth_running)
MIN_UPLIFT <- sens_dt[worth_running == TRUE & cost == C_OUTREACH, min(uplift)]
PROG_BREAKEVEN <- MIN_UPLIFT

ggplot(sens_dt, aes(x = factor(cost), y = factor(uplift))) +
  geom_tile(aes(fill = net), colour = pal$surface, linewidth = 1.6) +
  geom_text(aes(label = ifelse(worth_running, gbp_k(net), "do not run")),
            family = "", size = 2.75,
            colour = ifelse(sens_dt$net > 0.55 * max(sens_dt$net), "white", pal$ink)) +
  scale_fill_gradientn(colours = pal$seq[c(1, 2, 4, 6, 7)], labels = function(x) gbp_k(x),
                       name = "Annual net benefit  ") +
  guides(fill = guide_colourbar(barwidth = 11, barheight = 0.5, title.vjust = 0.92)) +
  scale_x_discrete(labels = function(x) gbp(as.numeric(x), 0)) +
  scale_y_discrete(labels = function(x) pc(as.numeric(x), 0)) +
  labs(title = "The programme only pays off in part of this range",
       subtitle = paste0("Base case (", pc(UPLIFT), " behavioural effect at ", gbp(C_OUTREACH, 2),
                         " per call) returns ", gbp(BASE_CELL), " a year"),
       x = "Cost per patient contacted", y = "Share of contacted patients who change behaviour",
       caption = "Darker means more value. Every cell re-solves for its own best threshold.") +
  theme_clinic(grid = "none", base_size = 11) +
  theme(axis.line.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "top")
```

Read this chart as a stress test rather than a forecast. Holding the call cost at
`r gbp(C_OUTREACH, 2)`, tripling it to `r gbp(6, 0)` would remove a large share of the value; on the
other axis, the programme needs the behavioural effect to reach at least
`r pc(MIN_UPLIFT, 0)` before it is worth running at all, and `r sum(!sens_dt$worth_running)` of the
`r nrow(sens_dt)` combinations tested say the same thing: do not run it. That is exactly why the
recommendation below treats the targeted-outreach play as something to pilot on a small scale first,
not something to roll out on faith.

# Recommendations

Four changes are recommended, in the order they should be made. The first three need no new evidence
because they either cost nothing or fix a pure process gap; the fourth should be tested on a small
group of patients before it is scaled network-wide.

```{r}
#| label: value-calc
#| include: false

n_sub72   <- nrow(d[sms_reminder_sent == 1 & reminder_hours_before < 72])
p1_saved  <- abs(eff_72) * n_sub72
p1_value  <- p1_saved * mean_net
p1_cost   <- 0

n_norem   <- nrow(d[sms_reminder_sent == 0])
gap_rem   <- mean(d[sms_reminder_sent == 0]$no_show) - mean(d[reminder_lead == "72h"]$no_show)
p2_saved  <- max(gap_rem, 0) * n_norem
p2_value  <- p2_saved * mean_net
p2_cost   <- n_norem * 0.06

p3_value  <- best_thr$benefit * scale_up
p3_cost   <- best_thr$cost * scale_up
p3_saved  <- UPLIFT * best_thr$tp * scale_up

n_cxl_ns  <- nrow(d[reminder_response == "Cancelled" & no_show == 1])
p4_recov  <- 0.5
p4_value  <- n_cxl_ns * p4_recov * mean(d[reminder_response == "Cancelled" & no_show == 1]$estimated_revenue_loss_gbp)
p4_cost   <- 2500
p4_saved  <- n_cxl_ns * p4_recov

plays <- data.table(
  Play = c("1. Default the reminder to 72 hours ahead",
           "2. Extend reminder coverage to everyone",
           "3. Targeted outreach on the flagged tail (pilot first)",
           "4. Auto-release slots when a patient cancels"),
  `Annual value` = c(gbp(p1_value), gbp(p2_value), gbp(p3_value), gbp(p4_value)),
  `Annual cost` = c(gbp(p1_cost), gbp(p2_cost), gbp(p3_cost), gbp(p4_cost)),
  `Net annual benefit` = c(gbp(p1_value - p1_cost), gbp(p2_value - p2_cost),
                           gbp(p3_value - p3_cost), gbp(p4_value - p4_cost))
)
p1_net <- p1_value - p1_cost; p2_net <- p2_value - p2_cost
p3_net <- p3_value - p3_cost; p4_net <- p4_value - p4_cost
total_net      <- p1_net + p2_net + p3_net + p4_net
now_net        <- p1_net + p2_net + p4_net
gated_net      <- p3_net
recovery_share <- total_net / net_loss
now_share      <- now_net / net_loss
gated_share    <- gated_net / net_loss
```

```{r}
#| label: exhibit-plays
#| echo: false
tbl(plays, caption = paste0("Exhibit 7. The recommended action plan. Combined it is worth ", gbp(total_net),
                            " a year, ", pc(recovery_share), " of the current net loss. ", gbp(now_net),
                            " of that (", pc(now_share), ") needs no pilot and could start next quarter."))
```

Play 3, targeted outreach, is the only one gated behind testing, because its value depends entirely on
an assumption (that a call changes `r pc(UPLIFT)` of contacted patients' behaviour) that nothing in
this dataset can confirm. Before it is scaled network-wide, it should be run as a genuine randomised
trial: call half of the flagged, highest-risk appointments and leave the other half as a comparison
group, then measure whether the called group actually attends more. Since only
`r pc(prec)` of flagged appointments are genuine no-shows, the programme needs to move the needle by
roughly `r pc(PROG_BREAKEVEN, 0)` to break even, a real but not enormous bar to clear, and one worth
testing cheaply before committing to it at scale.

# Limitations

Four caveats should travel with every number above. The dataset behind this report has the
statistical fingerprints of simulated rather than real-world data (no missing values, exact
arithmetic relationships between fields), so the analytical method is sound and reusable on a real
extract, but the specific pounds and percentages should be treated as illustrative rather than as
this network's confirmed budget line. Two fields meant to capture each patient's history of missed
appointments are empty in this extract, so no "chronically unreliable patient" segment could be
identified, only risk at the level of the individual appointment. The gap between patients who
confirm and patients who ignore a reminder mixes a real effect with patients simply self-selecting,
so it is used for ranking risk, never presented as a promised saving. And the whole business case
rests on one figure this dataset cannot supply: how much a phone call actually changes a patient's
behaviour, which is exactly what the recommended pilot is designed to measure before real money is
committed at scale.

# Conclusion

This network loses `r gbp(net_loss)` a year to appointments that are booked, staffed and never
attended, `r gbp(loss_per_appt, 2)` on every appointment it takes. That loss is driven by how the
clinic schedules and reminds, not by who its patients are: booking lead time, reminder timing and
unreleased cancellations explain most of what can be explained, and all three are within the clinic's
own control. A model built from those factors cannot say which specific patient will miss their
appointment, but it can rank appointments by risk well enough to make a targeted phone programme
profitable, and the fact that a far more complex model could not beat it is itself reassuring: the
relationship is simple enough to trust and cheap enough to maintain.

Put together, the four recommended changes are worth `r gbp(total_net)` a year, `r pc(recovery_share)`
of the current loss. The more important number is the split: `r gbp(now_net)`
(`r pc(now_share)`) needs no pilot and no new evidence, only a decision to change three operational
settings, while the remaining `r gbp(gated_net)` is real money on the table conditional on a small,
cheap trial confirming that outreach calls actually change behaviour. That is a deliberately modest
claim next to the size of the problem, and it is the honest one: this report shows the clinic exactly
how much of its no-show loss it can recover starting next quarter, and exactly what it needs to test
before promising the rest.

---

*Reproducibility: random seed `r SEED`, one value governing every step above. The full technical
report behind this summary, patient-level data audit, four-model bake-off with tuning detail,
calibration diagnostics and 41 supporting exhibits, is available as a companion document.*
